---
title: "Diversity"
author: "J. Ignacio Lucas Lledó"
date: "18/5/2020"
output: html_document
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE)
DA <- read.csv('../2020-02-27/DiversityAbundance.csv', header=TRUE, row.names=1)
DA$isoline <- as.factor(DA$isoline)
load('../2020-04-22/Diversity.RData')
library(ggplot2)
library(gridExtra)
library(plyr)
library(vegan)
```

On `2020-02-27` I created the CVS table `DiversityAbundance.cvs`, to share wit Zahida.
It includes diversity estimates (Shannon and Simpson) for every sample, in addition to
some other variables. That's all the information Zahida had to explore the relationship
between diversity indices and age. The abundance data from which diversity indices were
estimated had been filtered by mean early or late abundance, and by early or late prevalence.
Only ASVs present in at least 10 samples either early or late were included. While this
filter may be adequate for some purposes, it may not for diversity estimates. In
addition, there is a big difference between averaging diversity indices among replicates
and to directly estimate the diversity index of the joint replicates. The latter could be
more representative of an isoline and age.

On `2020-04-22`, I also estimated diversity indices, including phylogenetic diversity,
and using a much softer filter, applied separately for early and late samples: minimum
prevalence of 3 and minimum average relative abundance of $10^{-6}$.

After loading both the `DiversityAbundance.csv` data (as `DA`) and the diveristy data
from `2020-04-22` (dataframes `DiverIso.wide`, `DiverIso.long`, and `DiverRep.long`),
I combine the information of both analysis in two new data frames: `Rep` and `Iso`,
both in long format (with age as a factor). Letter *Z* identifies variables from the
`DiversityAbundance.cvs` file that I shared with Zahida, and letter *I* identifies
variables from the `2020-04-22` data set.

```{r comparison, message=FALSE}
# Assuming (after I checked) that the order of samples is the same in DA and
# in DiverRep.long.
Rep <- data.frame(
  Z.shannon = DA$shannon,
  I.shannon = DiverRep.long$Shannon,
  Z.simpson = DA$simpson,
  I.simpson = DiverRep.long$Simpson,
  isoline   = DA$isoline,
  age       = DA$age,
  seqrun    = DA$seqrun
)
p1 <- ggplot(Rep, aes(x=Z.shannon, y=I.shannon)) + geom_point() + geom_smooth(method='lm') +
  xlab('Minimum prevalence of 10') + ylab('Minimum prevalence of 3') + ggtitle('Shannon')
p2 <- ggplot(Rep, aes(x=Z.simpson, y=I.simpson)) + geom_point() + geom_smooth(method='lm') +
  xlab('Minimum prevalence of 10') + ylab('Minimum prevalence of 3') + ggtitle('Simpson')
grid.arrange(p1,p2,nrow=1)
```

Once checked that the data from `DiversityAbundance.csv` is mostly equivalent to 
the data produced later, on `2020-04-22` with a more relaxed threshold, I can
proceed. The following step is to produce isoline and age specific summary indices,
in a long format. I will take the average among replicates, to compare them to
the `DiverIso.long` values, which are computed directly from isoline and age specific
abundance data.

```{r Iso, message=FALSE}
Iso <- ddply(DA, .(age, isoline), summarize, Z.shannon = mean(shannon), Z.simpson = mean(simpson))
stopifnot(all.equal(Iso$age, DiverIso.long$age),
          all.equal(as.character(Iso$isoline), as.character(DiverIso.long$isoline)))
Iso$I.shannon <- DiverIso.long$Shannon
Iso$I.simpson <- DiverIso.long$Simpson
p1 <- ggplot(Iso, aes(x=Z.shannon, y=I.shannon)) + geom_point() + geom_smooth(method='lm') +
  xlab('Average among replicates, Prev. > 9') + ylab('Actual index, Prev. > 2') + ggtitle('Shannon')
p2 <- ggplot(Iso, aes(x=Z.simpson, y=I.simpson)) + geom_point() + geom_smooth(method='lm') +
  xlab('Average among replicates, Prev. > 9') + ylab('Actual index, Prev. > 2') + ggtitle('Simpson')
grid.arrange(p1, p2, nrow=1)
```

Close enough. Let's work on the `DiversityAbundance.csv` data, and try to replicate 
Zahida's results. She claims that Shannon index decreased with age, and reports an
$F_{1,23} = 7.3943$, and $p=0.0122$. Those numbers are the same as in Pau's report, who
claimed it was Simpson's, not Shannon's, the index significantly related with age.
In any case, I cannot reproduce those results.

The issue cannot be settled using only the data from `DiversityAbundance.cvs`. And because
the diversity indices depend on the filtering of rare ASVs, I think it's better to use
the relationship between diversity and minimum proportion to compare early and late samples,
in a way similar to rarefaction curves. Because the comparison of interest is between early
and late ages, I can just add up all samples in the same age to represent the total diversity
early and late.

```{r rarefaction}
TotEA.prop <- colSums(EA) / sum(EA)
TotLA.prop <- colSums(LA) / sum(LA)
rarefaction.early <- data.frame(
  breaks = sort(unique(round(c(TotEA.prop, TotLA.prop), 4))),
  age = factor('Early', levels=c('Early','Late')),
  Richness = NA,
  Simpson = NA,
  Shannon = NA
)
rarefaction.late <- data.frame(
  breaks = rarefaction.early$breaks,
  age = factor('Late', levels=c('Early', 'Late')),
  Richness = NA,
  Simpson = NA,
  Shannon = NA
)

for (i in 1:length(rarefaction.early$breaks)) {
  earlyFilter <- TotEA.prop >= rarefaction.early[i,'breaks']
  lateFilter  <- TotLA.prop >= rarefaction.early[i,'breaks'] # same as rarefaction.late[i,'breaks']
  rarefaction.early[i, 'Richness'] <- sum(earlyFilter)
  rarefaction.late[i, 'Richness']  <- sum(lateFilter)
  rarefaction.early[i, 'Shannon'] <- diversity(TotEA.prop[earlyFilter], index='shannon')
  rarefaction.late[i, 'Shannon'] <- diversity(TotLA.prop[lateFilter], index='shannon')
  rarefaction.early[i, 'Simpson'] <- diversity(TotEA.prop[earlyFilter], index='simpson')
  rarefaction.late[i, 'Simpson'] <- diversity(TotLA.prop[lateFilter], index='simpson')
}

rarefaction <- rbind(rarefaction.early, rarefaction.late)
rm(rarefaction.early, rarefaction.late)

ggplot(rarefaction[-1,], aes(x=breaks, y=Richness, color=age)) + geom_line() +
  xlab('Minimum frequency') + ylab('Amplicon richness') + 
  scale_x_log10() + scale_x_reverse() + scale_y_log10()

ggplot(rarefaction, aes(x=breaks, y=Shannon, color=age)) + geom_line() +
  xlab('Minimum frequency') + ylab('Shannon index') + scale_x_reverse()
ggsave(filename = 'Shannon_global.png')

ggplot(rarefaction, aes(x=breaks, y=Simpson, color=age)) + geom_line() +
  xlab('Minimum frequency') + ylab('Simpson index') + xlim(0.02, 0)

rm(rarefaction, TotEA.prop, TotLA.prop)
```

All three plots tell the same story. The Shannon index may be the clearest one.
The horizontal axis is reversed, to make the curve resemble a rarefaction curve:
the more to the right, the larger the number of ASVs counted. It could represent
an increasing sampling effort, ever finding rarer *species*.

What about individual isolines? I need a dataframe with columns: isoline, age,
shannon, simpson, richness... and minimum proportion. Let's try.

```{r perIsoline, fig.width=10, fig.height=10}
breaks <- sort(unique(round(c(EA.prop, LA.prop),4)))[-1]
diver.early <- do.call(rbind, lapply(breaks, function(x, M=EA.prop) {
  data.frame(minimum = rep(x, dim(M)[1]),
             isoline = as.factor(rownames(M)),
             age = factor('Early', levels=c('Early','Late')),
             richness = rowSums(M >= x),
             shannon  = apply(M, 1, function(y, MinFreq=x) diversity(y[y >= MinFreq], index='shannon')),
             simpson  = apply(M, 1, function(y, MinFreq=x) diversity(y[y >= MinFreq], index='simpson'))
  )
}))
diver.late <- do.call(rbind, lapply(breaks, function(x, M=LA.prop) {
  data.frame(minimum = rep(x, dim(M)[1]),
             isoline = as.factor(rownames(M)),
             age = factor('Late', levels = c('Early', 'Late')),
             richness = rowSums(M >= x),
             shannon = apply(M, 1, function(y, MinFreq=x) diversity(y[y >= MinFreq], index='shannon')),
             simpson = apply(M, 1, function(y, MinFreq=x) diversity(y[y >= MinFreq], index='simpson'))
  )
}))
diver <- rbind(diver.early, diver.late)
diver$isoline <- relevel(diver$isoline, '6')
ggplot(diver, aes(x=minimum, y=shannon, color=age)) + geom_line() +
  facet_wrap(~isoline) + xlab('Minimum frequency') +
  ylab('Shannon index') + scale_x_log10() + xlim(0.02, 0.00001) + theme(axis.text.x = element_text(angle=90))
ggsave(filename='Shannon_isoline.png')

ggplot(diver, aes(x=minimum, y=simpson, color=age)) + geom_line() +
  facet_wrap(~isoline) + xlab('Minimum frequency') +
  ylab('Shannon index') + scale_x_log10() + scale_x_reverse()

ggplot(diver, aes(x=minimum, y=richness, color=age)) + geom_line() +
  facet_wrap(~isoline) + xlab('Minimum frequency') +
  ylab('Shannon index') + scale_x_log10() + scale_y_log10() + scale_x_reverse()
```

It is clear that we cannot claim that diversity drops with age, at all. The opposite
could be claimed in many isolines, and also on average, unless we focused on the most
abundant ASVs and ignored the vast majority of low-frequency ASVs. In any case, the
lack of a clear pattern among isolines seems at odds with other studies and with our
own expectation to see a decline in diversity. After all, it is also evident (see results
from 2020-04-28 [here](https://htmlpreview.github.io/?https://github.com/IgnasiLucas/fly/blob/master/results/2020-04-28/figures.html))
that *Acetobacter* becomes the dominant genus, which seems to imply a marked reduction
of diversity. The reason for this discrepancy is that I am using ASVs, not genera, as
the sampling unit. Any two ASVs, whether they differ in one or a dozen nucleotides,
are considered different *species*. Whereas grouping ASVs in genera removes a lot of
that variation. That is why phylogenetic distances have such a big impact in the patterns
of diversity (see results from `2020-04-22` [here](https://htmlpreview.github.io/?https://github.com/IgnasiLucas/fly/blob/master/results/2020-04-22/diversity.html)).

Let's see what happens if I group the ASVs by genus.

```{r genusDiver, message=FALSE}
# I remove amplicons without genus assigned, but I don't filter amplicons by abundance.
filter <- ! is.na(Traits$Genus)
stopifnot(all.equal(rownames(Traits), colnames(Ar.prop)))
z <- as.data.frame(t(Ar.prop[, filter]))
Ar.genus <- do.call(rbind, lapply(split(z, Traits[filter, 'Genus']), colSums))
Ar.genus <- t(Ar.genus)
dim(Ar.genus)

# Although I didn't keep rownames in DiverRep.long, samples are in the same order as
# in SampleData, and Ar.genus. I add genus-wise diversity indices to DiverRep.long.

DiverRep.long$Shannon.Genus <- diversity(Ar.genus, index='shannon')
DiverRep.long$Simpson.Genus <- diversity(Ar.genus, index='shannon')
p1 <- ggplot(DiverRep.long, aes(x=Shannon.Genus, y=Shannon)) + geom_point() +
  scale_x_log10() + geom_smooth(method='lm')
p2 <- ggplot(DiverRep.long, aes(x=Shannon.Genus, y=PD)) + geom_point() +
  scale_x_log10() + geom_smooth(method='lm')
grid.arrange(p1,p2, nrow=1)

p1 <- ggplot(DiverRep.long, aes(x=age, y=Shannon.Genus)) + geom_boxplot(notch=TRUE) +
  scale_y_log10() + xlab('Age') + ylab('Shannon index') + ggtitle('Genus diversity')
p2 <- ggplot(DiverRep.long, aes(x=age, y=Shannon)) + geom_boxplot(notch=TRUE) +
  scale_y_log10() + xlab('Age') + ylab('Shannon index') + ggtitle('ASV diversity')
p3 <- ggplot(DiverRep.long, aes(x=age, y=Simpson.Genus)) + geom_boxplot(notch=TRUE) +
  scale_y_log10() + xlab('Age') + ylab('Simpson index') + ggtitle('Genus diversity')
p4 <- ggplot(DiverRep.long, aes(x=age, y=Simpson)) + geom_boxplot(notch=TRUE) +
  scale_y_log10() + xlab('Age') + ylab('Simpson index') + ggtitle('ASV diversity')
p5 <- grid.arrange(p2, p1, p4, p3, nrow=2)
p5
ggsave(filename='DiversityBoxplots.png', p5)
```

