---
title: "Summary of results of the analysis of gut microbiome data"
author: "J. Ignacio Lucas Lledó"
date: "25/4/2020"
output:
  pdf_document:
    fig_caption: yes
bibliography: "summary.bib"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
load('../2020-04-22/Diversity.RData')
```

# Pre-processing

Sequence data from the gut microbiome came in two batches. In all there samples from
24 isolines, and two time points considered *early* and *late* in life ages. There are
between 2 and 4 replicates of each isoline $\times$ age combination. Early and late
samples are balanced between the two batches. Most isolines got all their samples sequenced
in one of the two batches, except isolines 22, 23, and 24.

The sequencing center had already merged two forward and reverse reads from every sample,
with overall statistics showing good quality data. The only issue with the data was an
acute imbalance of sequencing effort among samples in the second sequencing batch. As a
result, early samples from isolines 27 and 35 had a relatively small number of sequences
(see figure \ref{fig:seqeff}).

```{r imbalance, echo=FALSE, fig.cap="\\label{fig:seqeff}Sequencing effort per isoline and sequencing batch"}
SampleData$SeqEff <- rowSums(Ar)
ggplot(SampleData, aes(x=isoline, y=SeqEff, fill=age)) +
  geom_bar(stat='identity') + facet_wrap(~seqrun) +
  xlab('Isoline') + ylab('Number of sequences') + theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

I used the package `dada2` [@Callahan2016] for *denoising*. This package implements
algorithmns to infer the exact sequences of different amplicons present in a sample,
to single nucleotide precision, despite sequencing errors. Thus, there is no need to
cluster similar sequences together in *operational taxonomic units* of doubtful
functional homogeneity. Instead, the unit of microbiome composition is the *amplicon
sequence variant* (ASV). 

The length distribution of ASVs was quite wide, most of its width occupied by noise
coming from spurious hybridization of sequencing primers with nuclear DNA. After some
manual BLAST searches, I retained only ASVs of lengths between 432 and 464. In all,
there were `r dim(taxa)[1]` different ASVs left.

See full report [here](https://htmlpreview.github.io/?https://github.com/IgnasiLucas/fly/blob/master/results/2019-12-12/denoising.html)

# Taxonomic attribution

The same package `dada2` allows for taxonomic attribution, based on public databases
of RNA sequences with known taxonomy. I removed a few additional ASVs with missing high level
taxonomic labels, or with spurious `Eurkaryotic` attribution (16S rRNA does not exist in
Eukaryotes). Only `r sum(! is.na(taxa[,'Species']))` (`r round(sum(! is.na(taxa[,'Species']))/dim(taxa)[1], 3)` %)
of ASVs got the species assigned. But `r round(sum(! is.na(taxa[,'Genus']))/dim(taxa)[1], 3)` got their genus assigned. 

```{r taxonomy, include=FALSE, fig.cap="\\label{fig:taxonomy}Lowest taxonomic level attributed."}
taxa <- as.data.frame(taxa)
maxLevel <- c('Nothing', names(taxa))
taxa$Level <- factor(maxLevel[rowSums(! is.na(taxa[,1:7])) + 1],
                   levels = c('Nothing','Kingdom','Phylum','Class',
                              'Order','Family','Genus','Species'),
                   ordered=TRUE)
ggplot(taxa, aes(x=1, fill=Level)) + geom_bar() + xlim(0,2) +
  xlab('')
```

I did not check the quality of the taxonomic attribution. The whole report
is available [here](https://htmlpreview.github.io/?https://github.com/IgnasiLucas/fly/blob/master/results/2020-02-11/taxonomy.html).

# Exploratory analysis

I run some exploratory analyses, including ordination plots, comparisons
of diversity levels between early and late samples, or relationship between
average proportion of an ASV in a sample and its prevalence among samples.
These results can be checked [here](https://htmlpreview.github.io/?https://github.com/IgnasiLucas/fly/blob/master/results/2020-02-27/phyloseq.html), and [here](https://htmlpreview.github.io/?https://github.com/IgnasiLucas/fly/blob/master/results/2020-03-25/ordination.html).

I was lucky to find a copy of *Numerical ecology* [@Legendre2012] avaiable online,
and I took the chance to learn about ordination methods and multivariate statistics.
I highly recommend the book. Know that it's the first time I use these methods, and
I apologize if there are mistakes.

My conclusions on the exploration of data are the following:

* **There is a batch effect**. Some ordination methods sparate clearly samples
sequenced in the first batch from those sequenced in the second. After
talking with Zahida, I understand the sequencing run itself may not be the
only source of variation, because samples sequenced later had also been
processed later, with potentiall different room temperature, or whatever.
Nothing to worry about, I think, but it's good to keep in mind.

* **Microbiome composition changes with age**. In later analyses (see below),
it becomes clear that some measures of diversity (not all) are lower in
late microbiomes.

# Differential abbundance

To test for a difference in abundance between early and late
samples requires specific packages, because we need to take into account
the biological variation among replicates of the same isoline. There are
thousands of ASVs and few replicates. The package I used is `DESeq2`
[@Love2014], originaly designed for RNA-seq data. 

Keeping false discovery rate below 0.001, I find `r dim(sigtab)[1]` ASVs
with significantly different abundances between early and late samples. Figure
\ref{fig:deseq2} shows their fold change, and how they are distributed
taxonomically.

```{r deseq, echo=FALSE, fig.cap="\\label{fig:deseq2}Fold change and taxonomic distribution of the amplicons with most differentiated abundances between early and late samples."}
#scale_fill_discrete <- function(palname = "Set1", ...) {
#    scale_fill_brewer(palette = palname, ...)
#}
# Phylum order
x = tapply(sigtab$log2FoldChange, sigtab$Phylum, function(x) max(x))
x = sort(x, TRUE)
#sigtab$Phylum = factor(as.character(sigtab$Phylum), levels=names(x))
# Genus order
x = tapply(sigtab$log2FoldChange, sigtab$Genus, function(x) max(x))
x = sort(x, TRUE)
#sigtab$Genus = factor(as.character(sigtab$Genus), levels=names(x))
ggplot(sigtab, aes(x=Genus, y=log2FoldChange, color=Phylum)) + geom_point(size=6) + 
  theme(axis.text.x = element_text(angle = -90, hjust = 0, vjust=0.5))
```

Remarkably, ASVs from four different phyla and at least 15 genera reduce their
abundances in late relative to early samples, while amplicons from only two
genera increase their abundances. Some of these amplicons (maybe the Cyanobacteria
and the Acetobacter with negative fold changes) could be false positives. But
the plot suggests a reduction in diversity with age.

# Redundancy analysis

The redundancy analysis was motivated by Pau Carazo's joint analysis of life
history data and abundance summary data. The summary data included diversity
measures and ordination vectors from a multidimensional scaling (MDS). The
MDS was based on a binary (presence/absence) distance measure among samples.
That distance was selected because it separated well early from late samples,
as well as samples from first and second batches. Even though there were some
interesting correlations, it was impossible to interpret the ordination axes
in terms of ASVs, because only a distance matrix and not the original abundances
entered the algorithm.

Redundancy analysis (RDA) seemed a good choice, because it is a multivariate method to
explain the variance in a set of variables with another set of variables. But it
has an important limitation: the number of explanatory variables cannot be larger
than the number of observations. The high dimensionality problem is typical, and
one way to deal with it is to run a PCA before the redundancy analysis [@Song2016].
Reducing the dimensionality of abundance data by selecting the first few principal
components would make interpretation difficult, but not impossible.

The main issue with the RDA is that we are bound to find false positives if we
do not test the global fit before believing the correlations that the RDA
finds [@Blanchet2008]. My first attempts were naive: the abundance data, early
and late in life, have a lot of variation that we do not expect to be correlated
with life-history traits variation. The RDA looks for the axes of variation that
maximize correlation, and it should not surprise us that it always finds something.

In the most recent attempts, I tried to use more meaningful predictable variables.
Zahida told me to use diversity measures, which I had forgotten before. I also
summarized the amplicon abundances by genus, and even by class. I focused on the
amplicons (and genera) with significantly different abundances between early and
late. I used abundances only early, or only late... Nothing that I tried produced
a significant global fit. The lowest p values I've seen are >0.11. You can see
[here](https://htmlpreview.github.io/?https://github.com/IgnasiLucas/fly/blob/master/results/2020-04-22/diversity.html) the full report of the latest attempt.

# References
