---
title: "Interpretation of RDA results"
author: "J. Ignacio Lucas Lledó"
date: "21/4/2020"
output: html_document
bibliography: interpretation.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(mice)
library(ggplot2)
library(vegan)
load('../2020-04-13/RDA.RData')
ls()
```

Recall the contents of the variables just imported from `2020-04-13`:

* `A` is the relative abundances of 614 amplicons in early life, and 418 amplicons
late in life. Columns are amplicons $\times$ age combinations, and rows correspond
to the 24 isolines with microbiome data. Only amplicons present in at least 5 isolines,
and with an average relative frequency of at least 5.0e-5 are included.
* `A.log` are the log transformed relative abundances. The logarithmic transformation
proved useful to reduce the skewness of abundance distributions across isolines.
Amplicons the abundance of which was still too skewed were removed, leaving 434
amplicons early and 318 amplicons late in life.
* `A.pca` is the principal component analysis of `A.log`. Because it is based on only
24 observations, only 23 eigenvalues are positive, the rest being zero.
* `batch` is a vector indicating the proportion of samples from each isoline processed
in the second (of two) batch. Used to account for technical variation.
* `LH` is the complete matrix of life-history traits for 29 isolines, including some
for which we don't have microbiome data.
* `LH.imp` is the object that holds the 10 imputations of the 2 missing values in
variable `ActuarialB`.

```{r}
barplot(cumsum(A.pca$sdev^2)[1:24] / sum(A.pca$sdev^2),
        main = 'PCA of log-transformed relative amplicon abundances',
        ylab = 'Accumulated portion of variance')
# LH.all is a list of LH matrices with imputed values.
LH.all <- complete(LH.imp, action='all')
validIsolines  <- row.names(A)
validVariables <- c('CSearly', 'CSSlope', 'EarlyRS', 'Rsen', 'AvLF', 'ActuarialB')
LH.all <- lapply(LH.all, function(x, f1=validIsolines, f2=validVariables) x[f1, f2])
```

The RDA analysis could take up to 22 components from the `A.pca`, because I also
include the batch effect. All together, 23: one less than the number of observations.
However, when testing for significance of the model, it seems that I need some degrees
of freedom, and the permutation test does not produce a *p* value unless there are
less than 22 components. In any case, the `rda()` function will only produces 6
eigenvectors, because there are 6 variables left in the matrix of life-history traits.

So, the question is what components of the variation in amplicon abundances carry any
correlation signal with life-history variables. Recall that PCA components are ordered
by the amount of variance they represent, which does not have to coincide with the amount
of correlation. Thus, any component could be correlated with life-history traits.

If we include all the components together, and test the global significance of the
model, the result is quite negative:

```{r globalTest}
RDA21 <- lapply(LH.all, function(lh) rda(lh ~ . + Condition(batch),
                                         data = as.data.frame(A.pca$scores[,1:21])))
RDA21.anova <- lapply(RDA21, anova)
RDA21.anova[[1]]
sapply(RDA21.anova, function(x) x$'Pr(>F)'[1])
```

The *p* values from all imputed data sets are non-significant. We should conclude
that there is no significant correlation (that we can detect) between gut microbiome
composition and life-history traits. @Blanchet2008 warn any results we get beyond
this point are bound to be false positives.

Well, but let's not stop the fun here. Keeping the warning in mind, let's see what
a positive result would look like.

```{r postTruth}

```

