---
title: "Redundancy analysis"
author: "J. Ignacio Lucas Lledó"
date: "13/4/2020"
output: html_document
bibliography: RDA.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r setup2, warning=FALSE, message=FALSE}
library(ggplot2)
library(gridExtra)
library(vegan)
library(Biostrings)
library(plyr)
library(e1071)
library(mice)
load('../2020-02-11/taxonomy.RData')
LH <- read.table('../../data/LifeHistoryVar.txt', header=TRUE, row.names=1)
ls()
```

# Underdetermination

**Redundancy analysis** was chosen because it does exactly what we need. It can tell
what portion of variance in a set of response variables is explained by another
set of variables. It is a combination of multivariate regression and principal
component analysis. It is like a PCA of the objects described by the response
variables, but *constrained* in such a way that the axis are chosen not to accumulate
most of the variance, but most of the correlation with the set of explanatory
variables. The resulting ordination maximizes the amount of variance in **Y** that
is explained by **X**.

But it has a limitation. The matrix of explanatory variables **X** is supposed to
have more cases or observations (in our case, **isolines**) than variables. That
is of course not the case. Our explanatory variables are the several hundreds of
amplicons identified in the gut microbiomes. This high-dimensionality problem is
very common when using XX century statistics with XXI century data. Mathematically
it is known as **underdetermination**. A linear system is considered underdetermined
when it has more unknowns than equations. 

In linear regression the equations are the  different samples or observations and
the *unknowns* are the regression coefficients corresponding to each independent
variable and the intercept. Typically (last century) a linear regression was an
*overdetermined* system: with more equations than unknowns. In that case, ordinary
least squares can be used to find the approximate solution, or the exact one if
it existed. Underdetermined systems seem harder to deal with. They have
either no solution or infinitely many.

So, in the best case, when regressing life-history traits against the abundances
of hundreds of amplicons, we would obtain an infinitude of possible combinations of
coefficients. In general, it seems that spurious correlations would appear [@Pezeshki2004].
But in practice, the implementation may not even bother. That is the case with the
`rda()` function in the `vegan` package: it does not use all the variables available,
but just the few first ones.

# Canonical correlation analysis

An alternative to redundancy analysis is canonical correlation analysis (CCorA).
The difference is analogous to that between regression and correlation. In CCorA
the two matrices **X** and **Y** are considered in a symmetrical way, without the
distinction between response and explanatory variables. CCorA finds *linear combinations*
of **X** and **Y** that have maximum correlation with each other. It seems to me
that CCorA avoids the underdetermination by reducing the dimensionality of both
**X** and **Y**. The canonical correlations are correlations between pairs
of linear combinations of variables in **X** and **Y**. There aren't any
multiple regressions involving the variables directly.

However, CCorA has a different problem. Because of the way canonical correlations
are computed, the covariance matrices of **X** and **Y** must be invertible. When
there are fewer observations than variables, the sample covariance matrix is not
full rank [@Legendre2012, p. 138], and therefore it is not invertible. I suspect
there should be ways to run CCorA that do not depend on the inversion of sample
covariance matrices. But I don't know of any such implementation.

# The strategy

An alternative is to reduce the dimensionality of the amplicon abundance
data before running the RDA. This is not ideal because the ordination of isolines based
on amplicon abundances will not maximize correlation with the life-histroy traits, but
just variance. Thus, if I run a PCA, the main axes will carry most of the variance but
not necessarily most of the correlation. This approach will only be successful if the
correlation signal is among the most variable amplicons, to be picked by the reduced
number of dimensions. Even asuming that the existing correlations involve amplicons
variable enough for that variation to be included in the first few axes of a PCA, an
important question is how many principal components we should use in the RDA. It is
tempting to use as many as possible to make sure that any correlation signal is still
present. But, including too many components would very easily generate spurious, false
positive correlations, as explained in @Song2016. This is a model-selection problem,
and it is related to the question of how many correlated signals there are [@Song2016].
It seems reasonable to assume that the number of correlated signals is small, even
compared with the number of isolines.

A PCA would be adequate, in my view, despite of the warnings usually raised in ecological
datasets. @Legendre2012 discourages the use of PCA on species abundance data when there
are many double zeros. The reason is that PCA preserves Euclidean distance between
sites, which counts double absences as contributing to similarity. As discussed in
`2020-03-25`, when using the abundance data as the independent variables, rather than
the dependent, it is acceptable for double absences to contribute to similarity, because
they are causaly equivalent.

About the high dimensionality problem, it does not seem to be fatal for PCA, as long as
we are willing to use fewer than 24 components, which is the number of isolines.

# Abundance data

Instead of importing the prepared dataset from `2020-03-25`, I will prepare the
matrices now again, to make this folder more self-contained. I start with the
abundance data matrix. Recall I take the taxonomy information (`taxa`) and the
abundance (`SeqTabNoChim`) from `2020-02-11`.

```{r abundanceData}
# Renaming the amplicons
dna <- DNAStringSet(row.names(taxa))
names(dna) <- sprintf("A%04i", seq(dim(taxa)[1])) # Naming all available sequences, even filtered
row.names(taxa)        <- names(dna)[match(row.names(taxa), dna)] 
colnames(SeqTabNoChim) <- names(dna)[match(colnames(SeqTabNoChim), dna)]
SampleData <- data.frame(
  age = factor(substr(row.names(SeqTabNoChim), 1, 1), levels=c('E','L'), labels=c('Early','Late')),
  isoline = factor(stringr::str_extract(rownames(SeqTabNoChim), "[[:digit:]]+"),
                   levels=as.character(c(6,10,11,12,14,15,17,19,20,22,23,24,25,26,27,28,29,30,31,33,35,36,38,39)),
                   ordered=FALSE),
  replicate = gsub("([EL][[:digit:]]+|.fastq.gz)", "", rownames(SeqTabNoChim)),
  seqrun = factor(1, levels=c(1,2))
)
run2 <- row.names(SeqTabNoChim) %in% c('E22C.fastq.gz', 'E23A.fastq.gz', 'E25A.fastq.gz',
                                       'E25B.fastq.gz', 'E25C.fastq.gz', 'E25D.fastq.gz',
                                       'E26A.fastq.gz', 'E26B.fastq.gz', 'E26C.fastq.gz',
                                       'E26D.fastq.gz', 'E27A.fastq.gz', 'E27B.fastq.gz',
                                       'E27C.fastq.gz', 'E28A.fastq.gz', 'E28B.fastq.gz',
                                       'E28C.fastq.gz', 'E28D.fastq.gz', 'E29A.fastq.gz',
                                       'E29B.fastq.gz', 'E29C.fastq.gz', 'E29D.fastq.gz',
                                       'E30A.fastq.gz', 'E30B.fastq.gz', 'E30C.fastq.gz',
                                       'E30D.fastq.gz', 'E31A.fastq.gz', 'E31B.fastq.gz',
                                       'E31C.fastq.gz', 'E31D.fastq.gz', 'E33A.fastq.gz',
                                       'E33B.fastq.gz', 'E33C.fastq.gz', 'E33D.fastq.gz',
                                       'E35A.fastq.gz', 'E35B.fastq.gz', 'E35C.fastq.gz',
                                       'E35D.fastq.gz', 'E36A.fastq.gz', 'E36B.fastq.gz',
                                       'E36C.fastq.gz', 'E36D.fastq.gz', 'E38A.fastq.gz',
                                       'E38B.fastq.gz', 'E38C.fastq.gz', 'E38D.fastq.gz',
                                       'E39A.fastq.gz', 'E39B.fastq.gz', 'E39C.fastq.gz',
                                       'E39D.fastq.gz', 'L23B.fastq.gz', 'L24C.fastq.gz',
                                       'L25A.fastq.gz', 'L25B.fastq.gz', 'L25C.fastq.gz',
                                       'L25D.fastq.gz', 'L26A.fastq.gz', 'L26B.fastq.gz',
                                       'L26C.fastq.gz', 'L26D.fastq.gz', 'L27A.fastq.gz',
                                       'L27B.fastq.gz', 'L27C.fastq.gz', 'L28A.fastq.gz',
                                       'L28B.fastq.gz', 'L28C.fastq.gz', 'L28D.fastq.gz',
                                       'L29A.fastq.gz', 'L29B.fastq.gz', 'L29C.fastq.gz',
                                       'L29D.fastq.gz', 'L30A.fastq.gz', 'L30B.fastq.gz',
                                       'L30C.fastq.gz', 'L30D.fastq.gz', 'L31A.fastq.gz',
                                       'L31B.fastq.gz', 'L31C.fastq.gz', 'L31D.fastq.gz',
                                       'L33A.fastq.gz', 'L33B.fastq.gz', 'L33C.fastq.gz',
                                       'L33D.fastq.gz', 'L35B.fastq.gz', 'L35C.fastq.gz',
                                       'L35D.fastq.gz', 'L36A.fastq.gz', 'L36B.fastq.gz',
                                       'L36C.fastq.gz', 'L36D.fastq.gz', 'L38A.fastq.gz',
                                       'L38B.fastq.gz', 'L38C.fastq.gz', 'L38D.fastq.gz',
                                       'L39A.fastq.gz', 'L39B.fastq.gz', 'L39C.fastq.gz',
                                       'L39D.fastq.gz')
SampleData$seqrun[run2] <- 2
row.names(SampleData) <- row.names(SeqTabNoChim)

isEarly <- SampleData$age == 'Early'
isLate  <- SampleData$age == 'Late'
EarlyAbundance <- SeqTabNoChim[isEarly,]
LateAbundance  <- SeqTabNoChim[isLate,]
EarlyIsoline <- SampleData[isEarly, 'isoline']
LateIsoline  <- SampleData[isLate,  'isoline']
# I found it impossible to do the following with plyr:
EA <- do.call(rbind, lapply(split(as.data.frame(EarlyAbundance), EarlyIsoline), colSums))
LA <- do.call(rbind, lapply(split(as.data.frame(LateAbundance),  LateIsoline),  colSums))
rm(EarlyAbundance, LateAbundance, EarlyIsoline, LateIsoline, isEarly, isLate, run2, SeqTabNoChim)
```

Now, `EA` and `LA` are matrices of *isoline* abundances early and late in life. That is,
I added up abundances in replicates of the same isoline. I want to use all isolines, even
though isolines 27 and 35 have very low counts among early samples. I will have to apply
a strict filter on amplicons, to exclude those not abundant enough to get sampled in (early)
isolines 27 and 35. In any case, removing low-frequency amplicons will help reduce the
dimensionality of the data.

I prepare the filters before transforming the data.

```{r filters, fig.width=10, warning=FALSE}
# It would not make sense to estimate proportions after filtering.
# I transpose to keep amplicons in columns and isolines in rows
EA.prop <- t(apply(EA, 1, function(x) x / sum(x)))
LA.prop <- t(apply(LA, 1, function(x) x / sum(x)))

AbundanceSummary <- data.frame(
  EA.meanProp = apply(EA.prop, 2, mean),
  LA.meanProp = apply(LA.prop, 2, mean),
  EA.prevalence = colSums(EA > 0),
  LA.prevalence = colSums(LA > 0)
)
p1 <- ggplot(AbundanceSummary, aes(x=EA.meanProp, y=EA.prevalence)) +
  geom_point(size=0.5) + scale_x_log10() + ggtitle('Early') +
  geom_vline(xintercept=5.0e-5, color='red') + geom_hline(yintercept=5, color='red')
p2 <- ggplot(AbundanceSummary, aes(x=LA.meanProp, y=LA.prevalence)) +
  geom_point(size=0.5) + scale_x_log10() + ggtitle('Late') +
  geom_vline(xintercept=5.0e-5, color='red') + geom_hline(yintercept=5, color='red')
grid.arrange(p1, p2, nrow=1)

EA.filter <- AbundanceSummary$EA.meanProp >= 5.0e-5 & AbundanceSummary$EA.prevalence >= 5
LA.filter <- AbundanceSummary$LA.meanProp >= 5.0e-5 & AbundanceSummary$LA.prevalence >= 5
```

Abundance data needs to be transformed because it is extremely skewed. In an accompanying
report, `transformations.Rmd` I decide that the most adequate transformation is logarithmic,
with an added constant equal to 1.0e-6. After transformation, the dataset will be further
reduced by removing several amplicons the abundance distribution of which are too skewed
after transformation.

```{r transformation}
colnames(EA.prop) <- paste0('E', colnames(EA.prop))
colnames(LA.prop) <- paste0('L', colnames(LA.prop))
A <- cbind(EA.prop[,EA.filter], LA.prop[,LA.filter])
A.log <- log10(A + 1.0e-6)
A.log.skewness <- apply(A.log, 2, skewness)
A.filter <- abs(A.log.skewness) <= 1.0
A.log <- as.data.frame(A.log[, A.filter])
dim(A.log)
```

# Principal Component Analysis

```{r PCA}
# I pre-calculate the covariance matrix to prevent princomp() from complaining
# that there are not enough observations.
A.log.covariance <- cov.wt(A.log)
A.pca <- princomp(A.log, covmat = A.log.covariance, scores = TRUE)
```

Now, I can use the scores of the isolines on a subset of the principal components
as the abundance data in an RDA. The question is how many components to use, always
less than 24.

# Covariate: sequencing run

The sequencing run is a sample's feature, not an isoline's. However, it has a
batch effect in microbiome composition. To remove such technical effect, I create
below a vector with the proportion of samples from the second batch in each isoline.
Most isolines belong entirely to one batch. Only three isolines differ in the proportion
of batch 2 samples between early and late time points. To simplify, I will consider
only the overall (not age-specific) proportion of second batch samples per isoline.

```{r batchEffect}
batch <- sapply(split(SampleData, SampleData$isoline),
                function(x) sum(x$seqrun == 2) / length(x$seqrun))
# To make sure the order is the same as in abundance data:
batch <- batch[row.names(A)]
```

# Life history traits

First, I standardize the variables.

```{r LH}
LH <- apply(LH, 2, function(x) (x - mean(x, na.rm=TRUE)) / sd(x, na.rm=TRUE))
pairs(LH)
```

The isoline with a noticeable outlier in `F1quality` was not included in the gut microbiome
study, and therefore will be dropped later. Then the distributions of the interesting
variables will be much more acceptable.

Now, I need to impute the two missing values of `ActuarlialB`. A simple search online
tells me that the preferred option is to use multiple imputed values, to get a sense of
the uncertainty in the imputation. The function `mice` from the `MICE` package generates
a set of imputations. Let's run the RDA for the first of them, that I extract with the
`complete` function.

I use the complete dataset of life-history traits to impute the two missing values, and
then remove the isolines (rows) for which we don't have microbiome data.

```{r imputation, message=FALSE, warning=FALSE}
LH.imp <- mice(LH, m=10)
LH.imp.1 <- complete(LH.imp, 1)
row.names(LH.imp.1) <- row.names(LH)
LH.imp.1 <- LH.imp.1[row.names(A), ]
```

# Redundancy  analysis

I use first one imputation to run a complete analysis. Once I get a clear idea of
what results we are most interested in, I can loop over the several imputations
suggested by `mice` and summarize the alternative results.

Recall that the `rda()` function from the `vegan` package expects the left hand side
of the input formula (or the `X` matrix) to be community composition data (species
abundances), because most ecologists seem interested in explaining composition with
environmental data, instead of using composition data to explain something else.
That is why it will list the life-history traits under the *species* label.

Using the `batch` as a *condition*, I am effectively applying what they call **partial**
redudancy analysis (pRDA). The problem with this approach is that the *R2 adjusted*
cannot be computed.

At this point, I want to use the same life-history traits that Pau Carazo used
to run a PCA previously, because they are very well selected and yielded reasonable
results. The variables are six: early climbing speed (`CSearly`), climbing speed
slope with age (`CSSlope`), early reproductive fitness (`EarlyRS`), reproductive
senescence (`Rsen`), average life span (`AvLF`), and acceleration of mortality rate
or $\beta$ component of a Gompertz fit on survival curve (`ActuarialB`).

```{r rda1}
interesting <- c('CSearly', 'CSSlope', 'EarlyRS', 'Rsen', 'AvLF', 'ActuarialB')
RDA0 <- rda(LH.imp.1[, interesting] ~ Condition(batch),
            data = as.data.frame(A.pca$scores[,1:5]), scale=FALSE)
RDA5 <- rda(LH.imp.1[, interesting] ~ . + Condition(batch),
            data = as.data.frame(A.pca$scores[,1:5]), scale=FALSE)
```

```{r biplots, fig.width=10, fig.height=10}
par(mfrow=c(2,2))
  plot(RDA5, choices=c(1,2), display=c('sp','bp'), scaling=2)
  plot(RDA5, choices=c(3,2), display=c('sp','bp'), scaling=2)
  plot(0, type='n', axes=FALSE, ann=FALSE)
  plot(RDA5, choices=c(3,1), display=c('sp','bp'), scaling=2)
par(mfrow=c(1,1))
```

This is just an example. I used one possible imputation, and I decided to include 5
components from the PCA on amplicon abundance data. Note that I fitted two models,
one with 5 components and the other with none. I can compare them, to run a global
test, as recommended by @Blanchet2008:

```{r globalTest}
anova(RDA0, RDA5)
```

According to this test, the variation along the 5 principal components in amplicon
relative abundances among isolines does not really help explain the life-history
traits' variation. Before taking this result as definitive, I will save the data,
close this document and move to a different folder.

```{r sessionInfo}
save(A, A.log, A.pca, batch, LH, LH.imp, taxa, file = 'RDA.RData')
sessionInfo()
```
