---
title: "Transformations"
author: "J. Ignacio Lucas Lledó"
date: "19/4/2020"
output: html_document
bibliography: 'RDA.bib'
---

```{r setup, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(e1071)
library(Biostrings)
library(tidyr)
library(ggplot2)
library(plyr)
load('../2020-02-11/taxonomy.RData')
```

Here, I want to evaluate different transformations of abundance data, with the main
goal of reducing skewness. For a principal component analysis (PCA), it is not strictly
required to have multinormality, but skewness should be avoided [@Legendre2012, p. 411].

Below I prepare the two amplicon abundance matrices, `EA` and `LA`, for early and late
abundances respectively. The amplicons (columns) are given a short name, and replicates
of the same isoline are added, so that only 24 rows are left.

```{r data}
# I use the taxa matrix to name amplicons, to be consistent with previous scripts.
dna <- DNAStringSet(row.names(taxa))
names(dna) <- sprintf("A%04i", seq(dim(taxa)[1]))
row.names(taxa)        <- names(dna)[match(row.names(taxa), dna)] 
colnames(SeqTabNoChim) <- names(dna)[match(colnames(SeqTabNoChim), dna)]
SampleData <- data.frame(
  age = factor(substr(row.names(SeqTabNoChim), 1, 1), levels=c('E','L'), labels=c('Early','Late')),
  isoline = factor(stringr::str_extract(rownames(SeqTabNoChim), "[[:digit:]]+"),
                   levels=as.character(c(6,10,11,12,14,15,17,19,20,22,23,24,25,26,27,28,29,30,31,33,35,36,38,39)),
                   ordered=FALSE),
  replicate = gsub("([EL][[:digit:]]+|.fastq.gz)", "", rownames(SeqTabNoChim))
)
# Not using seqrun here.

row.names(SampleData) <- row.names(SeqTabNoChim)

isEarly <- SampleData$age == 'Early'
isLate  <- SampleData$age == 'Late'
EarlyAbundance <- SeqTabNoChim[isEarly,]
LateAbundance  <- SeqTabNoChim[isLate,]
EarlyIsoline <- SampleData[isEarly, 'isoline']
LateIsoline  <- SampleData[isLate,  'isoline']
# I found it impossible to do the following with plyr:
EA <- do.call(rbind, lapply(split(as.data.frame(EarlyAbundance), EarlyIsoline), colSums))
LA <- do.call(rbind, lapply(split(as.data.frame(LateAbundance),  LateIsoline),  colSums))
rm(EarlyAbundance, LateAbundance, EarlyIsoline, LateIsoline, isEarly, isLate, SeqTabNoChim)
```

Absolute abundances must be turned into relative abundances. Otherwise, abundances are not
dimensionally homogeneous across isolines, because of varying sequencing efforts. Then, I
can use prevalence and average abundance to filter amplicons. Removing the rarest amplicons
does not change the distribution of relative abundances of the remaining amplicons across
isolines, but may eliminate some of the most skewed amplicons.

```{r relative}
EA.prop <- t(apply(EA, 1, function(x) x / sum(x)))
LA.prop <- t(apply(LA, 1, function(x) x / sum(x)))

AbundanceSummary <- data.frame(
  EA.meanProp = apply(EA.prop, 2, mean),
  LA.meanProp = apply(LA.prop, 2, mean),
  EA.prevalence = colSums(EA > 0),
  LA.prevalence = colSums(LA > 0)
)

EA.filter <- AbundanceSummary$EA.meanProp >= 5.0e-5 & AbundanceSummary$EA.prevalence >= 5
LA.filter <- AbundanceSummary$LA.meanProp >= 5.0e-5 & AbundanceSummary$LA.prevalence >= 5

EA.high <- EA.prop[,EA.filter]
LA.high <- LA.prop[,LA.filter]

colnames(EA.high) <- paste0('E', colnames(EA.high))
colnames(LA.high) <- paste0('L', colnames(LA.high))
A <- cbind(EA.high, LA.high)

Skew <- data.frame(Amplicon   = substr(colnames(A), 2, 6),
                   Age        = substr(colnames(A), 1, 1),
                   meanProp   = NA,
                   Prevalence = NA,
                   Relative   = apply(A, 2, skewness),
                   stringsAsFactors = FALSE)
Skew[Skew$Age == 'E', 'meanProp'] <- AbundanceSummary[Skew[Skew$Age == 'E','Amplicon'], 'EA.meanProp']
Skew[Skew$Age == 'L', 'meanProp'] <- AbundanceSummary[Skew[Skew$Age == 'L','Amplicon'], 'LA.meanProp']
Skew[Skew$Age == 'E', 'Prevalence'] <- AbundanceSummary[Skew[Skew$Age == 'E', 'Amplicon'], 'EA.prevalence']
Skew[Skew$Age == 'L', 'Prevalence'] <- AbundanceSummary[Skew[Skew$Age == 'L', 'Amplicon'], 'LA.prevalence']
```

The metric chi-square transformation was one of the transformations suggested by
@Legendre2001 that produced some interesting non-Euclidean distances among original
vectors when calculating the Euclidean distances among transformed vectors. The reasons
to use such non-Euclidean distances I think had more to do with the double-zero
problem than with skewness. Actually, the metric chi-square transformation does not
reduce the skewness at all. What I need are non-linear transformations applied separately
to each amplicon's relative abundances.

```{r basicTransforms}
A.sqrt <- sqrt(A)
A.cube <- A ^ (1/3)
A.logA <- log10(A + 1.0e-4)
A.logB <- log10(A + 1.0e-5)
A.logC <- log10(A + 1.0e-6)
A.logD <- log10(A + 1.0e-7)
A.logE <- log10(A + 1.0e-8)
A.logF <- log10(A + 1.0e-9)

Skew$Sqrt <- apply(A.sqrt, 2, skewness)
Skew$Cube <- apply(A.cube, 2, skewness)
Skew$LogA <- apply(A.logA, 2, skewness)
Skew$LogB <- apply(A.logB, 2, skewness)
Skew$LogC <- apply(A.logC, 2, skewness)
Skew$LogD <- apply(A.logD, 2, skewness)
Skew$LogE <- apply(A.logE, 2, skewness)
Skew$LogF <- apply(A.logF, 2, skewness)

SkewLong <- pivot_longer(Skew, cols=5:13, names_to='Transform', values_to='Skewness')
SkewLong$Transform <- factor(SkewLong$Transform,
                             levels=c('Relative','Sqrt','Cube','LogA','LogB','LogC','LogD','LogE','LogF'),
                             ordered = TRUE)
ggplot(SkewLong, aes(x=Transform, y=Skewness)) + geom_boxplot()
```

The logarithmic transformation with an added constant of 1e-6 successfully reduces
the skewness for most amplions, but makes some amplicons much more (left) skewed
than they were before. Those are relatively prevalent amplicons, which did not even
require much of a transformation in the first place.

Here I have applied the same basic transformation to all descriptors, because all
of them are the same kind of data, and it makes interpretation easier. It is also
possible to apply a different transformation (or none) to different amplicons. It
could even make biological sense that some amplicons affect some life-history traits
in a linear fashion, while other have an exponential effect. However, I have no
*a priori* reason to treat different amplicons differently. After transformation,
only a few become too negatively skewed. Even though they are among the most abundant
amplicons, not all the abundant amplicons are negatively affected by the logarithmic
transformation. I think the most reasonable option is to remove from the analysis
the amplicons the abundance distribution of which are too skewed after transformation. 
Setting the thresholds at -1 and 1 would leave 434 amplicons in early life, and 318
in late life. 

```{r sessionInfo}
sessionInfo()
```
