---
title: "Barplots and phylogenetic trees"
author: "J. Ignacio Lucas Lledó"
date: "28/4/2020"
output: html_document
bibliography: figures.bib
notice: |
  @Wickham2016, @McMurdie2013, @Paradis2018, @Yu2017, @Yu2020
---

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(phyloseq)
library(ggplot2)
library(ape)
library(ggtree)
library(tidytree)
load('../2020-02-27/ps.RData')
load('../2020-04-22/Diversity.RData')
# These are amplicons significantly associated with low or high
# levels of PC1 and PC3 of life-history traits. See 2020-05-01
PC1low <- c('A0108','A0121','A0128','A0126','A0131','A0140',
            'A0150','A0152','A0144','A0166','A0171')
PC3low <- c('A0207','A0229','A0252','A0254','A0257','A0269',
            'A0264','A0261','A0260','A0281','A0282','A0128')
PC3high <- 'A0076'
```

# Barplot of abundances

The reason I have not produced a barplot of abundances before is twofold.
First, I didn't like the output of the `plot_bar()` function in the phyloseq
package: it doesn't make it easy at all to remove the black border around the
boxes. And second, the large number of amplicons and the low diversity of the
compositions make barplots not very informative, unless carefully customized.

I tried again the `plot_bar()` function in `phyloseq`, following this
[tutorial](https://joey711.github.io/phyloseq/plot_bar-examples.html). However,
I find it necessary to *melt* the dataset and work directly with `ggplot2`,
to remove the default borders around each amplicon's abundance.

In all barplots below, I stack relative abundances from the same isoline and age.
That makes the height of the bar equal to the number of replicates for each
isoline and age, which is either 3 or 4. In order to represent compositions
relative to the same reference level, I divide samples' relative abundances
by the number of samples from the same isoline and age. Thus, the height of
the bars represent average values among replicates of the same isoline and
age.

```{r barplot1, message=FALSE}
NumRep <- table(sample_data(ps)[,c('age', 'isoline')])
NumRep
ps.prop.melted <- psmelt(ps.prop)
ps.prop.melted$IsoAbundance <- numeric(length = length(ps.prop.melted$Abundance))
isEarly <- ps.prop.melted$age == 'Early'
isLate  <- ps.prop.melted$age == 'Late'
ps.prop.melted[isEarly, 'IsoAbundance'] <- ps.prop.melted[isEarly, 'Abundance'] /
  NumRep['Early', ps.prop.melted[isEarly, 'isoline']]
ps.prop.melted[isLate, 'IsoAbundance'] <- ps.prop.melted[isLate, 'Abundance'] /
  NumRep['Late', ps.prop.melted[isLate, 'isoline']]
rm(isEarly, isLate)
ggplot(ps.prop.melted, aes(x = isoline, y = IsoAbundance, fill = Class)) +
  geom_bar(stat = 'identity', position = 'stack') +
  facet_wrap(~ age) +
  theme(axis.text.x = element_text(angle = 90)) +
  ylab('Relative abundance')
ggsave(filename = 'Barplot_Isoline_Abundance_Class.png')
```

I found the Class taxonomic level to be the easiest to visualize in the barplot
above. There are 7 Classes and 6 Phyla. The correspondence is the following:

```{r table, echo=FALSE}
unique(tax_table(ps)[,c('Phylum','Class')])
```

In the barplot above, some bars do not reach the maximum height of 1. The reason is
that low-abundance amplicons, filtered out in a previous step made up the missing
fraction.

Below, I break down the average relative abundances by Class.

```{r barplot3}
z <- split(ps.prop.melted, ps.prop.melted$Class)
names(z)
p <- list(length=length(z))
for (i in 1:length(z)) {
  z[[i]]$Genus <- droplevels(z[[i]]$Genus)
}
ggplot(z$Bacteroidia, aes(x=isoline, y=IsoAbundance, fill=Genus)) +
  geom_bar(stat='identity', position='stack') +
  facet_wrap(~ age) +
  theme(axis.text.x=element_text(angle=90)) +
  ylab('Relative abundance') + ggtitle('Bacteroidetes')
ggsave(filename = 'Barplot_Isoline_Abundance_Bacteroidetes.png')

ggplot(z$Verrucomicrobiae, aes(x=isoline, y=IsoAbundance, fill=Genus)) +
  geom_bar(stat='identity', position='stack') +
  facet_wrap(~ age) +
  theme(axis.text.x=element_text(angle=90)) +
  ylab('Relative abundance') + ggtitle('Verrucomicrobia')
ggsave(filename = 'Barplot_Isoline_Abundance_Verrucomicrobia.png')

ggplot(z$Oxyphotobacteria, aes(x=isoline, y=IsoAbundance, fill=Order)) +
  geom_bar(stat='identity', position='stack') +
  facet_wrap(~ age) +
  theme(axis.text.x=element_text(angle=90)) +
  ylab('Relative abundance') + ggtitle('Cyanobacteria')
ggsave(filename = 'Barplot_Isoline_Abundance_Cyanobacteria.png')

ggplot(z$Actinobacteria, aes(x=isoline, y=IsoAbundance, fill=Genus)) +
  geom_bar(stat='identity', position='stack') +
  facet_wrap(~ age) +
  theme(axis.text.x=element_text(angle=90)) +
  ylab('Relative abundance') + ggtitle('Actinobacteria')
ggsave(filename = 'Barplot_Isoline_Abundance_Actinobacteria.png')

ggplot(z$Bacilli, aes(x=isoline, y=IsoAbundance, fill=Genus)) +
  geom_bar(stat='identity', position='stack') +
  facet_wrap(~ age) +
  theme(axis.text.x=element_text(angle=90)) +
  ylab('Relative abundance') + ggtitle('Firmicutes')
ggsave(filename = 'Barplot_Isoline_Abundance_Firmicutes.png')

ggplot(z$Gammaproteobacteria, aes(x=isoline, y=IsoAbundance, fill=Genus)) +
  geom_bar(stat='identity', position='stack') +
  facet_wrap(~ age) +
  theme(axis.text.x=element_text(angle=90)) +
  ylab('Relative abundance') + ggtitle('Proteobacteria, Gammaproteobacteria')
ggsave(filename = 'Barplot_Isoline_Abundance_Gammaproteobacteria.png')

ggplot(z$Alphaproteobacteria, aes(x=isoline, y=IsoAbundance, fill=Genus)) +
  geom_bar(stat='identity', position='stack') +
  facet_wrap(~ age) +
  theme(axis.text.x=element_text(angle=90)) +
  ylab('Relative abundance') + ggtitle('Proteobacteria, Alphaproteobacteria')
ggsave(filename = 'Barplot_Isoline_Abundance_Alphaproteobacteria.png')
```

In the plots above we can see clear patters. Remarkably, the disappearance
of almost anything that is not *Acetobacter*, except for *Lactobacillus*.
The reduction of within-isoline diversity (Alpha diversity?) was known
before. Now it is also clear that Beta diversity also goes down: it is **not**
the case that isolines differentiate with age. Rather, they converge to
a similar state, as in an ecological succession.

# Phylogenetic trees

The analysis with Lefse [@Segata2011] (`2020-05-01`) identified a few amplicons
significantly associated with low (only one with high) loadings
of isolines on principal components of life-history traits. All of
those amplicons belong to *Acetobacter*, which is not surprising, for
it is the most abundant genus. If amplicons aparently associated with
life-history traits were clustered within the *Acetobacter* phylogeny,
the possibility that they are actually rellevant for the fly's physiology
would be more plausible.


```{r ggtree}
# The amplicons in the tree are these (ordered):
AmpFilt <- sort(colnames(EA)[EA.filter | LA.filter])
stopifnot(all.equal(AmpFilt, names(fitGTR$data)))
TreeTraits <- Traits[AmpFilt,]
TreeAceto <- drop.tip(fitGTR$tree, TreeTraits[! TreeTraits$Genus %in% 'Acetobacter', 'label'])
TibbAceto <- as_tibble(TreeAceto)
TibbAceto$lefse <- character(length=dim(TibbAceto)[1])
TibbAceto[TibbAceto$label %in% PC1low, 'lefse'] <- 'PC1low'
TibbAceto[TibbAceto$label %in% PC3low, 'lefse'] <- 'PC3low'
TibbAceto[TibbAceto$label %in% PC3high, 'lefse'] <- 'PC3high'
TDAceto <- as.treedata(TibbAceto)
ggtree(TDAceto, aes(color=lefse)) + theme_tree2()
```

I need to remove the long branches. And the colors on leaves don't help.
I also discard the least significant amplicon
from the `PC3low` set, which belongs to a different clade, and messes up the plot.

```{r prunning}
TibbAceto[TibbAceto$branch.length > 0.4,]
# Manually identified these tips in long branches:
TDAcetoShort <- treeio::drop.tip(TDAceto, c('A1543', 'A1910', 'A2130', 'A2172',
                                      'A1818', 'A1067', 'A1306', 'A1787', 'A1962'))
ggtree(TDAcetoShort, layout = 'circular') +
  geom_hilight(node = MRCA(TDAcetoShort, PC1low), fill='green', alpha=0.5) +
  geom_hilight(node = MRCA(TDAcetoShort, PC3low[-12]), fill='blue', alpha=0.5) +
  geom_hilight(node = MRCA(TDAcetoShort, PC3high), fill='red', alpha=0.5) +
  geom_cladelabel(node = MRCA(TDAcetoShort, PC1low), label='Low PC1', offset.text=0.15) +
  geom_cladelabel(node = MRCA(TDAcetoShort, PC3low[-12]), label='Low PC3', offset.text=0.05) +
  geom_cladelabel(node = MRCA(TDAcetoShort, PC3high), label='High PC3', offset.text=0.1) +
  geom_treescale(x=0.7, offset=5)

ggsave(filename = 'Acetobacter_Circular_Lefse_highlight.png')

ggtree(TDAcetoShort, layout = 'rectangular') +
  geom_hilight(node = MRCA(TDAcetoShort, PC1low), fill='green', alpha=0.5) +
  geom_hilight(node = MRCA(TDAcetoShort, PC3low[-12]), fill='blue', alpha=0.5) +
  geom_hilight(node = MRCA(TDAcetoShort, PC3high), fill='red', alpha=0.5) +
  geom_cladelabel(node = MRCA(TDAcetoShort, PC1low), label='Low PC1', offset.text=0.05) +
  geom_cladelabel(node = MRCA(TDAcetoShort, PC3low[-12]), label='Low PC3', offset.text=0.05) +
  geom_cladelabel(node = MRCA(TDAcetoShort, PC3high), label='High PC3', offset.text=0.1) +
  geom_treescale(x=0.5, offset=5)

ggsave(filename = 'Acetobacter_Rectangular_Lefse_highlight.png')
```

After removing just one amplicon (the least significant one) from the set of
amplicons identified by Lefse as associated with low fitness loadings on PC3,
it becomes clear that significant amplicons are phylogenetically clustered.
Amplicons with the same presumed effect on fitness components are closely related
within the genus tree. In principle, this looks like a strong support for the
biological significance of the finding. It also suggests that had the amplicons
been grouped appropriately, the signal would have been stronger.

In the figure, not all highlighted amplicons are significant. The highlighted
clades are defined by the most recent common ancestor of significant amplicons
with the same presumed effect.

```{r sessionInfo}
save(TDAceto, TDAcetoShort, file='figures.RData')
sessionInfo()
```

# References
