---
title: "Taxonomy attribution"
author: "J. Ignacio Lucas Lledó"
date: "11/01/2020"
output: html_document
bibliography: taxonomy.bib
params:
  TABSEQDATA: '../2019-12-12/denoising.RData'
  SILVA_TRAINING: '../../data/silva_nr_v132_train_set.fa.gz'
  SILVA_SPECIES: '../../data/silva_species_assignment_v132.fa.gz'
  NUM_THREADS: 12
---

```{r setup, message=FALSE}
library(dada2)
library(RcppParallel)   # this is to set the number or threads.
TABSEQDATA     <- params$TABSEQDATA
SILVA_TRAINING <- params$SILVA_TRAINING
SILVA_SPECIES  <- params$SILVA_SPECIES
NUM_THREADS    <- as.numeric(params$NUM_THREADS)
```

# Taxonomy attribution

This step takes almost an hour. I use 'cache=TRUE' to save the partial result and speed
up subsequent runs if the chunk has not been modified. I am using the default `minBoot=50`,
and not asking for the bootstrap values. This is an implementation of the RDP Naive
Bayesian Classifier algorithm described in [@Wang2007], with kmer size 8 and 100 bootstrap
replicates. I am using the Silva taxonomic databases [@Callahan2018]. We need to cite these
references.

```{r taxa, cache=TRUE}
load(TABSEQDATA)
setThreadOptions(numThreads = NUM_THREADS)
taxa <- assignTaxonomy(SeqTabNoChim, SILVA_TRAINING, multithread=TRUE)
taxa <- addSpecies(taxa, SILVA_SPECIES)
dim(taxa)
```

# Summary of results
The `assignTaxonomy()` and `addSpecies()` functions produce a matrix of characters, with these
columns: `r paste(colnames(taxa), collapse=', ')`. The row names are the sequences themselves.
The first question I have is how successful the taxonomy attribution was. In the matrix, an
undetermined taxonomic level is an `NA` value. I just need to count `NA` values to know the
taxonomic level at which a sequence got annotated: 0 means the species is known; 1, the genus...
and 7, not even the kingdom (*nothing* is known).

```{r countNA}
success <- table(rowSums(is.na(taxa)))
row.names(success) <- c(rev(colnames(taxa)), 'Nothing')
success
```

So, most sequences got a genus assigned; only 16, the species, and so on. Maybe I can blast the
least known sequences manually to see what they are. What are the 2 sequences without a Kingdom
assigned? Are they Martian?

```{r Mars}
query <- row.names(taxa)[is.na(taxa[,'Kingdom'])]
cat(paste(c('>unknown1', '>unknown2'), query, sep='\n', collapse='\n'))
```

Both are sequences from the nuclear *Drosophila melanogaster* genome, amplified because of
inspecific hybridization of the primers. I should remove them from the `SeqTabNoChim` matrix
of abundances.

```{r remove}
SeqTabNoChim <- SeqTabNoChim[, !(colnames(SeqTabNoChim) %in% query)]
dim(SeqTabNoChim)
```

What about the 23 sequences without a Phylum?

```{r noPhylum}
query <- row.names(taxa)[is.na(taxa[,'Phylum']) & !(is.na(taxa[,'Kingdom']))]
table(taxa[query,'Kingdom'])
```
These seem to be eukaryotic sequences. Are there more?

```{r kingdoms}
table(taxa[,'Kingdom'])
```
Well, there is only one more presumably eukaryotic sequence, which must have been annotated further. Which one?

```{r eukaryota}
taxa[row.names(taxa)[taxa[,'Kingdom'] == 'Eukaryota' & !(is.na(taxa[,'Phylum']))],]
```

Well, eukaryotes do not have 16S rRNA. So all these must be either wrong taxonomic attributions or inspecific
amplifications from the *D. melanogaster* genome. I manually blasted the 24 "eukaryotic" sequences, and only
two or three could be real, bacterial, 16S rRNA that have been missclassified. The rest are Drosophila's.
I think the sensitive thing to do is to remove all them from `SeqTabNoChim`.

```{r removeEukaryotes}
wrongSeqs <- row.names(taxa[taxa[,'Kingdom'] %in% 'Eukaryota',])
SeqTabNoChim <- SeqTabNoChim[, !(colnames(SeqTabNoChim) %in% wrongSeqs)]
```

# Reformating of table of abundances
The matrix `SeqTabNoChim` is now `r dim(SeqTabNoChim)[1]` rows and `r dim(SeqTabNoChim)[2]` columns.
Row names are the names of the fastq files from each sample, like: `r paste(head(row.names(SeqTabNoChim)), collapse=', ')`.
And column names are the sequences. I want a very long data frame with these columns:

- Time: a factor with levels 'early' and 'late'.
- Isoline.
- Replicate: A, B, C...
- Phylum.
- Class.
- Order.
- Family.
- Genus.
- Species.
- Count: Number of reads of that species observed in that sample.
- Proportion: Fraction of reads in the sample corresponding to this species.
- Sequence: not sure if I should include it.



# References

```{r save}
save(c(taxa, SeqTabNoChim), file='taxonomy.RData')
sessionInfo()
```
