---
title: "Taxonomy attribution"
author: "J. Ignacio Lucas Lledó"
date: "11/01/2020"
output: html_document
bibliography: taxonomy.bib
params:
  TABSEQDATA: '../2019-12-12/denoising.RData'
  SILVA_TRAINING: '../../data/silva_nr_v132_train_set.fa.gz'
  SILVA_SPECIES: '../../data/silva_species_assignment_v132.fa.gz'
  NUM_THREADS: 12
---

```{r setup, message=FALSE}
library(dada2)
library(RcppParallel)   # this is to set the number or threads.
library(tidyr)
library(stringr)
library(plyr)
TABSEQDATA     <- params$TABSEQDATA
SILVA_TRAINING <- params$SILVA_TRAINING
SILVA_SPECIES  <- params$SILVA_SPECIES
NUM_THREADS    <- as.numeric(params$NUM_THREADS)
```

# Taxonomy attribution

This step takes almost an hour. I tried using 'cache=TRUE' to save the partial result and speed
up subsequent runs if the chunk has not been modified. But it does not work well. I am using the default `minBoot=50`, 
and not asking for the bootstrap values. This is an implementation of the RDP Naive
Bayesian Classifier algorithm described in [@Wang2007], with kmer size 8 and 100 bootstrap
replicates. I am using the Silva taxonomic databases [@Callahan2018]. We need to cite these
references.

```{r loading}
load(TABSEQDATA)
setThreadOptions(numThreads = NUM_THREADS)
```

```{r taxa}
taxa <- assignTaxonomy(SeqTabNoChim, SILVA_TRAINING, multithread=TRUE)
taxa <- addSpecies(taxa, SILVA_SPECIES)
dim(taxa)
```

# Summary of results
The `assignTaxonomy()` and `addSpecies()` functions produce a matrix of characters, with these
columns: `r paste(colnames(taxa), collapse=', ')`. The row names are the sequences themselves.
The first question I have is how successful the taxonomy attribution was. In the matrix, an
undetermined taxonomic level is an `NA` value. I just need to count `NA` values to know the
taxonomic level at which a sequence got annotated: 0 means the species is known; 1, the genus...
and 7, not even the kingdom (*nothing* is known).

```{r countNA}
success <- table(rowSums(is.na(taxa)))
names(success) <- c(rev(colnames(taxa)), 'Nothing')
success
```

So, most sequences got a genus assigned; only 16, the species, and so on. Maybe I can blast the
least known sequences manually to see what they are. What are the 2 sequences without a Kingdom
assigned? Are they Martian?

```{r Mars}
query <- row.names(taxa)[is.na(taxa[,'Kingdom'])]
cat(paste(c('>unknown1', '>unknown2'), query, sep='\n', collapse='\n'))
```

Both are sequences from the nuclear *Drosophila melanogaster* genome, amplified because of
inspecific hybridization of the primers. I should remove them from the `SeqTabNoChim` matrix
of abundances.

```{r remove}
SeqTabNoChim <- SeqTabNoChim[, !(colnames(SeqTabNoChim) %in% query)]
dim(SeqTabNoChim)
```

What about the sequences without a Phylum?

```{r noPhylum}
query <- row.names(taxa)[is.na(taxa[,'Phylum']) & !(is.na(taxa[,'Kingdom']))]
table(taxa[query,'Kingdom'])
```
These seem to be eukaryotic sequences. Are there more?

```{r kingdoms}
table(taxa[,'Kingdom'])
```

Eukaryotes do not have 16S rRNA. So all these must be either wrong taxonomic attributions or inspecific
amplifications from the *D. melanogaster* genome. I manually blasted the 24 "eukaryotic" sequences, and only
two or three could be real, bacterial, 16S rRNA that have been missclassified. The rest are Drosophila's.
I think the sensitive thing to do is to remove all them from `SeqTabNoChim`.

```{r removeEukaryotes}
wrongSeqs <- row.names(taxa[taxa[,'Kingdom'] %in% 'Eukaryota',])
SeqTabNoChim <- SeqTabNoChim[, !(colnames(SeqTabNoChim) %in% wrongSeqs)]
```

# Reformating of table of abundances
The matrix `SeqTabNoChim` is now `r dim(SeqTabNoChim)[1]` rows and `r dim(SeqTabNoChim)[2]` columns.
Row names are the names of the fastq files from each sample, like: `r paste(head(row.names(SeqTabNoChim)), collapse=', ')`.
And column names are the sequences. I want a very long data frame with these columns:

- Time: a factor with levels 'early' and 'late'.
- Isoline.
- Replicate: A, B, C...
- Phylum.
- Class.
- Order.
- Family.
- Genus.
- Species.
- Count: Number of reads of that species observed in that sample.
- Proportion: Fraction of reads in the sample corresponding to this species.
- Sequence: not sure if I should include it.

```{r reformat}
TA <- as.data.frame(SeqTabNoChim)
TA$fastq <- row.names(TA)
TA <- pivot_longer(TA, 1:(dim(TA)[2] - 1), names_to='sequence', values_to='abundance')
TA$time <- factor('early', levels=c('early','late'), ordered=TRUE)
TA[startsWith(TA$fastq, 'L'), 'time'] <- 'late'
table(TA$time)
TA$isoline <- str_extract(TA$fastq, "[[:digit:]]+")
table(TA$isoline)
TA$replicate <- gsub("([EL][[:digit:]]+|.fastq.gz)", "", TA$fastq)
table(TA$replicate)
TA$Phylum <- factor(taxa[TA$sequence, 'Phylum'], levels = unique(taxa[,'Phylum']))
TA$Class  <- factor(taxa[TA$sequence, 'Class'],  levels = unique(taxa[,'Class']))
TA$Order  <- factor(taxa[TA$sequence, 'Order'],  levels = unique(taxa[,'Order']))
TA$Family <- factor(taxa[TA$sequence, 'Family'], levels = unique(taxa[,'Family']))
TA$Genus  <- factor(taxa[TA$sequence, 'Genus'],  levels = unique(taxa[,'Genus']))
TA$Species <- factor(taxa[TA$sequence, 'Species'], levels = unique(taxa[,'Species']))
# Totals is the total number of sequence counts from a sample. 
Totals <- rowSums(SeqTabNoChim)
TA$total  <- Totals[TA$fastq]
TA$proportion <- TA$abundance / TA$total
```

In data frame `TA` (for *taxonomy and abundance*), `total` is the total number of counts
of all sequences in the same fastq file (or sample). Anf `proportion` is the proportion
of those reads corresponding to a sequence. Next, I want to pool replicates of the same
isoline and sampling time. Then, there are two ways to estimate the proportion of a sequence
among the replicates. I can average the previously estimated proportions among replicates,
or I can pool the sequence abundances and the totals (sequencing efforts) of all replicates
and calculate the proportion again. The result would be the same if the average is weighted
by totals, as it should be. The question is *what's the error of such an average proportion?*.
We probably need a negative binomial model or similar. Time to check what people out there do
and switch folders.

```{r summarise, echo=FALSE, eval=FALSE}
TA2 <- ddply(TA, .(sequence,time,isoline), summarize,
             Phylum = Phylum[1],
             Class = Class[1],
             Order = Order[1],
             Family = Family[1],
             Genus = Genus[1],
             Species = Species[1],
             repliNum = length(replicate),
             pooledCount = sum(abundance),
             pooledTotal = sum(Total),
             pooledProp = sum(abundance)/sum(Total), # equal to meanProp?
             meanProp = weighted.mean(proportion, abundance),
             sdProp = sd(proportion))
```

# References

```{r save}
save(taxa, SeqTabNoChim, file='taxonomy.RData')
sessionInfo()
```
