---
title: "Taxonomy attribution"
author: "J. Ignacio Lucas Lledó"
date: "11/01/2020"
output: html_document
bibliography: taxonomy.bib
params:
  TABSEQDATA: '../2019-12-12/denoising.RData'
  SILVA_TRAINING: '../../data/silva_nr99_v138_train_set.fa.gz'
  SILVA_SPECIES: '../../data/silva_species_assignment_v138.fa.gz'
  NUM_THREADS: 12
---

```{r setup, message=FALSE}
library(dada2)
library(RcppParallel)   # this is to set the number or threads.
library(tidyr)
library(stringr)
library(plyr)
TABSEQDATA     <- params$TABSEQDATA
SILVA_TRAINING <- params$SILVA_TRAINING
SILVA_SPECIES  <- params$SILVA_SPECIES
NUM_THREADS    <- as.numeric(params$NUM_THREADS)
```

This step takes almost an hour. I tried using 'cache=TRUE' to save the partial result and speed
up subsequent runs if the chunk has not been modified. But it does not work well. I am using
`minBoot=80`, as suggested by the `dada2` tutorial,
and not asking for the bootstrap values. This is an implementation of the RDP Naive
Bayesian Classifier algorithm described in [@Wang2007], with kmer size 8 and 100 bootstrap
replicates. I am using the Silva taxonomic databases [@Callahan2018; @Quast2012; @Yilmaz2013].
We need to cite these references.

```{r loading}
load(TABSEQDATA)
setThreadOptions(numThreads = NUM_THREADS)
```

```{r taxa}
taxa <- assignTaxonomy(SeqTabNoChim, SILVA_TRAINING, multithread=TRUE, minBoot=80)
taxa <- addSpecies(taxa, SILVA_SPECIES, allowMultiple=TRUE)
dim(taxa)
```

# Summary of results
The `assignTaxonomy()` and `addSpecies()` functions produce a matrix of characters, with these
columns: `r paste(colnames(taxa), collapse=', ')`. The row names are the sequences themselves.
The first question I have is how successful the taxonomy attribution was. In the matrix, an
undetermined taxonomic level is an `NA` value. I just need to count `NA` values to know the
taxonomic level at which a sequence got annotated: 0 means the species is known; 1, the genus...
and 7, not even the kingdom (*nothing* is known).

```{r countNA}
success <- table(rowSums(is.na(taxa)))
names(success) <- c(rev(colnames(taxa)), 'Nothing')
success
```

So, most sequences got a genus assigned; only `r success[1]`, the species, and so on. 
Maybe I can blast the least known sequences manually to see what they are. What are
the `r success['Nothing'] sequences without a Kingdom assigned? Are they Martian?

```{r Mars}
query <- row.names(taxa)[is.na(taxa[,'Kingdom'])]
cat(paste(sprintf('>unknown%i', 1:success[8]), query, sep='\n', collapse='\n'))
```

The last time I manually BLASTed those sequences, all three of them were found to be
most similar to some strain of *Acetobacter persici*. I ignore the reason why the
Silva database fails to attribute even a Kingdom to these sequences. Because BLAST
suggests they are valid bacterial sequences, I do not exclude them.

Let's take a look at the distribution of Kingdoms to make sure that everything is
Bacterial:

```{r}
table(taxa[,'Kingdom'])
```

The only sequences I am willing to exclude from the analysis on the grounds of
their taxonomy attribution are those assigned to the Mitochondria family, within
the order Rickettsailes. There are `r sum(taxa[,'Family'] %in% 'Mitochondria')`
mitochondrial sequences. 

```{r removeMitochondrial}
# Recall, I use "%in%" instead of "==" because NA values in 'Family' cannot
# be said to be equal or not to anything, be they can be said to be members
# or not of a set.
wrongSeqs <- row.names(taxa[taxa[,'Family'] %in% 'Mitochondria',])
SeqTabNoChim <- SeqTabNoChim[, !(colnames(SeqTabNoChim) %in% wrongSeqs)]
dim(SeqTabNoChim)
```

# References

```{r save}
save(taxa, SeqTabNoChim, file='taxonomy.RData')
sessionInfo()
```
