---
title: "Ordination methods"
author: "J. Ignacio Lucas Lledó"
date: "25/3/2020"
output: html_document
bibliography: ordination.bib
---

```{r setup1, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r setup2, warning=FALSE, message=FALSE}
library(phyloseq)
library(ggplot2)
library(gridExtra)
library(ape)
library(vegan)
load('../2020-02-27/ps.RData')
load('../2020-02-11/taxonomy.RData')
LH <- read.table('../../data/LifeHistoryVar.txt', header=TRUE)
ls()
head(LH)
```
# Introduction

On 2020-02-27, I run several ordination analyses, mostly following a tutorial, without
much knowledge of what I was doing. Results were heavily dependent on the distance
measure used, and also on the aggressiveness of the filtering step (not shown). A more
detailed analysis is necessary.

# Exploratory ordination analysis

## NMDS with Bray-Curtis distance

State of the art microbial ecology studies use Bray-Curtis distances and non-metric
multidimensional scaling (NMDS) for exploratory analysis [@PerezCobas2015; @Rosas2018].
NMDS is a numerical method (iterative, not analyitic). It is not based on eigenvectors
and it does not provide a unique solution. It has the advantage of using all the variation
in the data (not only that associated with the main axes) to distribute the objects
(samples) in a reduced-dimension space. In such a mapping, the ranks of dissimilarities
among pairs of objects are respected, if possible. NMDS is applied to a matrix of distances,
where the information on the original variables is collapsed: axes cannot be interpreted in
terms of the original variables [@Paliy2016].

The Bray-Curtis distance is a popular choice among ecologists [@Ricotta2017]. It is
computed from raw abundance data (rather than presence/absence data) and it does not
include double zeros in the comparison, preventing samples without common species from
looking very similar. In absolute terms, differences between abundant species contribute
the same to the Bray-Curtis distance as (absolute) differences between rare species
[@Legendre2012, p. 287]. But, because the same absolute difference is relatively smaller
in an abundant than in a rare species, it can also be said that a relative difference between
abundant species is weighted more heavily than the same relative difference among rare
species [@Ricotta2017, p. 203].

The excess of zeros in our dataset is not a concern for the Bray-Curtis distance, but the
large scale and differences in overall abundance among samples is. If left untransformed,
the few distances involving samples with extremely high or low coverage dominate the space,
to the point of making the plot irrelevant. Let's take a look at the outliers:

```{r outliers}
z <- sample_sums(ps)
summary(z)
z[z < 3000 | z > 300000]
rm(z)
```

Samples E31D and L35C have too few counts to make any valid inference of their composition.
They should be removed. Samples L29C and L29D can be normalized or downsampled, just like the
rest, to approximate an even sequencing coverage among samples. But, actually, after removing
the two samples with the fewest counts, Bray-Curtis distance does a very good job even without
any transformation. However, sample sizes should be standardized, to avoid ordination reflect
primarly sample size.

```{r BrayCurtisNMDS, fig.width=10}
ps.pruned <- prune_samples(sample_sums(ps) > 3000, ps)
medianSize <- median(sample_sums(ps.pruned))
ps.pruned <- transformSampleCounts(ps.pruned, function(x, t=medianSize) round(t * x / sum(x)))
ord.nmds.bray <- ordinate(ps.pruned, method="NMDS", distance="bray",
                          trymax=1000, trace=0, autotransform=FALSE)
ord.nmds.bray
p1 <- plot_ordination(ps, ord.nmds.bray, color='age')
p2 <- plot_ordination(ps, ord.nmds.bray, color='seqrun')
grid.arrange(p1, p2, nrow=1, top='NMDS with Bray-Curtis distances')
```

The stress value is `r ord.nmds.bray$stress`. Values below 0.2 are just fine. Below 0.15
would be much better. In any case, the main conclusion I draw from this plot is that the
main difference between early and late microbiomes is that early ones are more dispersed.

I have tried to produce an NMDS using Sørensen's distance, which is the Bray-Curtis equivalent
for presence-absence data [@Tamas2001]. Unfortunately, the NMDS algorithm fails to converge
in any acceptable solution when using Sørensen's distance.

## Principal Coordinates Analysis (PCoA)

PCoA also uses a distance matrix as input data, rather than the raw abundance values. Its
purpose is also to display a set of objects in a reduced dimensionality space while respecting
as much as possible the original distances among them.

```{r PCoA, fig.width=10}
ord.pcoa.bray <- ordinate(ps.pruned, method='PCoA', distance='bray', correction='cailliez')
p1 <- plot_ordination(ps.pruned, ord.pcoa.bray, color='age')
p2 <- plot_ordination(ps.pruned, ord.pcoa.bray, color='seqrun')
grid.arrange(p1, p2, nrow=1, top='PCoA with Bray-Curtis distances')
```

Maybe the main difference between PCoA and NMDS is that in PCoA plot, only the variation
along two main axes is represented, while in NMDS all the variation is squeezed in the
two dimensions.

In PCoA, we can use any distance, like Sørensen's, which lets us evaluate the effect of
using binary (presence/absence) instead of quantitative data.

```{r sorensen, fig.width=10}
# Note that using 'sor' as the distance, the resulting ordination is completely wrong,
# because the 'sor' distance (from betadiver) returns similarities, not dissimilarities.
# And apparently, phyloseq's ordinate does not take that into account. Distance 't' is
# the complement of Sørensen's index.
ord.pcoa.sor <- ordinate(ps.pruned, method='PCoA', distance='t', correction='cailliez')
p1 <- plot_ordination(ps.pruned, ord.pcoa.sor, color='age')
p2 <- plot_ordination(ps.pruned, ord.pcoa.sor, color='seqrun')
grid.arrange(p1, p2, nrow=1, top='PCoA with Sørensen distance')
```
Neither Bray-Curtis nor Sørensen distances are metric: they do not follow the triangle
inequality axiom. Thus, they cannot be properly represented in a Euclidean space. Negative
eigenvalues appear. Nevertheless, the results are both very reasonable. It is actually
surprising that the binary Sørensen distance separate so well between early and late
samples, and between first and second batches.

Apparently, one of the original motivations for the development of PCoA was to incorporate
phylogenetic information in the distances among communities, something that principal
component analysis (PCA) cannot do. It does sound like a very good idea. Let's see how
the phylogeny contributes to the PCoA plot:

```{r unifrac, warning=FALSE, fig.width=10}
ord.pcoa.unifrac  <- ordinate(ps.pruned, method='PCoA', distance='unifrac')
ord.pcoa.wunifrac <- ordinate(ps.pruned, method='PCoA', distance='wunifrac')
p1 <- plot_ordination(ps.pruned, ord.pcoa.unifrac, color='age')
p2 <- plot_ordination(ps.pruned, ord.pcoa.unifrac, color='seqrun')
p3 <- plot_ordination(ps.pruned, ord.pcoa.wunifrac, color='age')
p4 <- plot_ordination(ps.pruned, ord.pcoa.wunifrac, color='seqrun')
grid.arrange(p1, p2, nrow=1, top='PCoA with unweighted unifrac distances')
grid.arrange(p3, p4, nrow=1, top='PCoA with weighted unifrac distances')
```

For some reason, the *wunifrac* distance gives unstable results. In any case, the point is that
both phylogenetic distances among amplicons and abundances have a strong influence in the
overall ordination of samples. But up to now, the most clear distinction between early and
late-life samples was achived with a binary distance.

## The role of double absences

Ecologists avoid distance measures in which common absences contribute to similarity.
This is called 'the double zero problem' [@Legendre2012, p. 253]:

> The double-zero problem shows up in ecology because of the special nature of
> species descriptors. Species are known to have unimodal distributions along
> environmental gradients (Whittaker, 1967). Hutchinson’s (1957) niche theory states
> that species have ecological preferences, meaning that they are more likely to be found
> at sites where they encounter appropriate living conditions; the distribution of a
> species has its mode at this optimum value. A species becomes rare and eventually
> absent as one departs from optimal conditions. If a species is present at two sites, this
> is an indication of the similarity of these sites; but if a species is absent from two sites,
> it may be because the two sites are both above the optimal niche value for that species,
> or both are below, or else one site is above and the other is below that value. One
> cannot tell which of these circumstances is the correct one.
>
> It is thus preferable to abstain from drawing any ecological conclusion from the
> absence of a species at two sites. In numerical terms, this means to skip double zeros
> altogether when computing similarity or distance coefficients using species presence-
> absence or abundance data. Coefficients of this type are called asymmetrical because
> they treat zeros in a different way than other values.

For gut microbial communities of a single host species in laboratory conditions,
it is hard to believe that we are sampling beyond the optimal environment of any microbial
species. In the current setting, we probably should not expect taxa to show a unimodal
distribution. Even across isolines, there is no *a priori* reason to think that some
isolines offer a better environment to particular bacteria than others.

In any case, the focus of the experimental design are not the causes but the consequences
of variation in gut microbiome composition. What is considered similar or different
depends on whether the variable is explanatory or a response. Because an absence can
be caused by several unknown factors (e.g., either too low or too high salinity), it
should not count as a similarity. But when assessing the consequences of an absence,
any two absences are causaly equivalent, and they should count in the similarity measure.

[@Faith1983] suggested a binary similarity measure that considers double absences as neutral,
contributing neither to similarity nor to distance. It is defined as $C=\frac{a + 0.5d}{N}$,
where $a$ is the number of species present in both sites, $d$ the number of double
absences, and $N$, the total number of species. If double absences were to count just
as much as a mismatch to distance, $C$ would turn into Russell and Rao similarity: $\frac{a}{N}$.
On the other side, if double absences were to contribute to similarity as much as double
presences, then $C$ would turn into the Raw similarity, $\frac{a+d}{N}$. They can
all be easily implemented with the function `designdist()`. Corresponding distances
are obtained subtracting similarities from 1.

```{r FaithBinary}
ord.pcoa.RussellRao <- pcoa(designdist(otu_table(ps.pruned),
                                       method = '1 - a / (a + b + c + d)',
                                       abcd = TRUE),
                            rn = sample_names(ps.pruned), correction = 'lingoes')
ord.pcoa.faith      <- pcoa(designdist(otu_table(ps.pruned),
                                       method = '1 - (a + 0.5 * d) / (a + b + c + d)',
                                       abcd = TRUE),
                            rn = sample_names(ps.pruned), correction = 'lingoes')
ord.pcoa.raw        <- pcoa(designdist(otu_table(ps.pruned),
                                       method = '1 - (a + d) / (a + b + c + d)',
                                       abcd = TRUE),
                            rn = sample_names(ps.pruned), correction = 'lingoes')
p1 <- plot_ordination(ps.pruned, ord.pcoa.RussellRao, color='age')
p2 <- plot_ordination(ps.pruned, ord.pcoa.RussellRao, color='seqrun')
p3 <- plot_ordination(ps.pruned, ord.pcoa.faith, color='age')
p4 <- plot_ordination(ps.pruned, ord.pcoa.faith, color='seqrun')
p5 <- plot_ordination(ps.pruned, ord.pcoa.raw, color='age')
p6 <- plot_ordination(ps.pruned, ord.pcoa.raw, color='seqrun')
grid.arrange(p1, p2, nrow=1, top='PCoA with Russell-Rao distance. Double absences as mismatches')
grid.arrange(p3, p4, nrow=1, top='PCoA with Faith distance. Double absences as neutral')
grid.arrange(p5, p6, nrow=1, top='PCoA with Raw distance. Double absences as double presences')
```

[@Tamas2001] generalize several distance measures that take absences into account from
the binary to the quantitative scale. Their genaralization of a binary distance defined
by D.P. Faith in 1983 can help us reveal the effect of doble absences in our data set.
The index is a similarity and is given by equation 9b in [@Tamas2001]:

$FA2q_{jk} = \frac{\sum_{i=1}^n min(x_{ij}, x_{ik}) + 0.5\sum_{i=1}^n (\max_j(x_{ij}) - \max(x_{ij},x_{ik}))}{\sum_{i=1}^n\max_j(x_{ij})}$

Unfortunately, the `designdist()` function does not allow a direct implementation of
this similarity index. I use my own implementation below. I make the 0.5 factor variable
to obtain quantitative analogs of Raw and Russell and Rao distances.

```{r Faith2Q, fig.width=10}
Faith2q <- function(x, W=0.5){
  maxj <- apply(x, 2, max)  # vector of maximum abundances species achieve across sites.
  d <- matrix(0, nrow=nrow(x), ncol=nrow(x))  # comparing rows, assumed to be 'sites'
  for (j in 2:nrow(x)) {
    for (i in 1:(j-1)) {
      d[j,i] <- 1 - 
        (sum(pmin.int(x[i,], x[j,])) + W * sum(maxj - pmax.int(x[i,], x[j,]))) / sum(maxj)
    }
  }
  return(as.dist(d))
}
ord.pcoa.faith2qW0 <- pcoa(Faith2q(otu_table(ps.pruned), W=0),
                         rn=sample_names(ps.pruned), correction='lingoes')
ord.pcoa.faith2qW05 <- pcoa(Faith2q(otu_table(ps.pruned), W=0.5),
                            rn=sample_names(ps.pruned), correction='lingoes')
ord.pcoa.faith2qW1 <- pcoa(Faith2q(otu_table(ps.pruned), W=1),
                           rn=sample_names(ps.pruned), correction='lingoes')
p1 <- plot_ordination(ps.pruned, ord.pcoa.faith2qW0, color='age')
p2 <- plot_ordination(ps.pruned, ord.pcoa.faith2qW0, color='seqrun')
p3 <- plot_ordination(ps.pruned, ord.pcoa.faith2qW05, color='age')
p4 <- plot_ordination(ps.pruned, ord.pcoa.faith2qW05, color='seqrun')
p5 <- plot_ordination(ps.pruned, ord.pcoa.faith2qW1, color='age')
p6 <- plot_ordination(ps.pruned, ord.pcoa.faith2qW1, color='seqrun')
grid.arrange(p1, p2, nrow=1, top='PCoA with quantitative Rusell-Rao distance. Double absences as mismatches')
grid.arrange(p3, p4, nrow=1, top='PCoA with quantitative Faith distance, with double absences neutral')
grid.arrange(p5, p6, nrow=1, top='PCoA with quantitative Raw distance, with double absences as double presences')
```

# Redundancy analysis

Ordination analyses are a set of multivariate methods that can be classified in three
main groups, according to their main purpose: exploratory, interpretative and
discriminatory methods [@Paliy2016]. 

In our model, microbiome composition is but a predictor variable of life-history
variables related to fitness and measured on isolines. To be more precise, for every
isoline we have measured two microbiome compositions (with replicates): one early and
one late in life. They are expected to be correlated, but different, and both may
contribute to explain life-history traits. The main question is to what extent microbiome
composition influences life-history traits, and how. This is not an exploratory
analysis, but an interpretative one, in which we use a large set of variables (taxa
abundances) to try to explain part of the variation in life-history traits.

A redundancy analysis (RDA in the literature) seems to be the best choice, although
multivariate versions of generalized linear models (GLM) are also a possibility [@Paliy2016].
The `vegan` package includes an `rda()` function, that takes community data as the
presumed response variables, and expects environmental data as the explanatory ones.
That's the typical usage in ecology, and also microbial ecology [@Zhang2012, e.g.].
But our goal is the reverse: to explain isoline (community-level) properties by their
microbiome composition. 

...





# Some quotes

 [@Legendre2001, p. 276]:

> If rare species are well sampled and truly rare, they
> may be used as indicators of the conditions that may exist
> at some sites only. In that case, they should receive
> high weight, as they do in CA and, to a lesser extent, in
> PCA after the chi-square transformation (Table 2). In
> contrast, when rare species are observed sporadically at
> some sites but, as the result of “sampling error”, not at
> others where they are also present, giving them high
> weight in the analysis is unwise. This phenomenon is ex-
> acerbated in environments where sampling is conducted
> blindly – for example, in aquatic and soil ecology. Co-
> efficients such as the Euclidean, chord, species profile or
> Hellinger distances do not give high weights to the rare
> species (Table 2).

Also from [@Legendre2001, p. 279]:

> For the analysis of community gradients, it does not matter
> that an analysis gives high weights to the rare species
> when the end-product is simply a reduced-space ordination diagram.
> [...] The weights given to rare species do matter, however, when the
> end-product is a test of significance of the relationship between
> species composition and a set of explanatory variables, or a test
> of factors in a multiple analysis-of-variance model following the
> db-RDA approach of Legendre and Anderson (1999).

# Session information

```{r sessionInfo}
sessionInfo()
```

# References
