---
title: "The role of gut microbiome diversity in life-history traits"
author: "J. Ignacio Lucas Lledó"
date: "22/4/2020"
output: html_document
bibliography: diversity.bib
---

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(picante)
library(vegan)
library(ggplot2)
library(gridExtra)
library(Biostrings)
library(DECIPHER)  # alignment
library(tidyr)
library(mice) # multiple imputation
library(lme4)
library(plyr)
library(e1071)
load('../2020-02-11/taxonomy.RData')
load('../2020-03-17/DESeq2.RData')
LH <- read.table('../../data/LifeHistoryVar.txt', header=TRUE, row.names=1)

```

# Introduction

Zahida encouraged me to include measures of diversity among the predictor
variables. I will take the chance to expand the description of diversity
estimates initated in `2020-02-27`. Back then I built a maximum likelihood
phylogenetic tree of the amplicons, following directions form @Callahan2016.
I want to use a tree to estimate phylogenetic diversities. But, because I
changed the naming of amplicons since then, I will build the tree again,
to make sure everything is compatible.

I loaded the taxonomy information from `2020-02-11`, which includes the
taxonomy assignment and the original abundance table, both using sequences
as names.

# Data preparation
## Abundance data
The idea is to relate diversity measures early and late in life with life-history
variables of isolines. Thus, it makes sense to collapse the different replicates of
the same isoline and time point. I add up those abundances, to estimate one diversity
index per isoline and age with all the information, instead of having to average a
few measures later. This is the same processing I did on `2020-04-13`.

```{r abundance}
dna <- DNAStringSet(row.names(taxa))
names(dna) <- sprintf("A%04i", seq(dim(taxa)[1])) # Naming all available sequences, even filtered
row.names(taxa)        <- names(dna)[match(row.names(taxa), dna)] 
colnames(SeqTabNoChim) <- names(dna)[match(colnames(SeqTabNoChim), dna)]
SampleData <- data.frame(
  age = factor(substr(row.names(SeqTabNoChim), 1, 1), levels=c('E','L'), labels=c('Early','Late')),
  isoline = factor(stringr::str_extract(rownames(SeqTabNoChim), "[[:digit:]]+"),
                   levels=as.character(c(6,10,11,12,14,15,17,19,20,22,23,24,
                                         25,26,27,28,29,30,31,33,35,36,38,39)),
                   ordered=FALSE),
  replicate = gsub("([EL][[:digit:]]+|.fastq.gz)", "", rownames(SeqTabNoChim)),
  seqrun = factor(1, levels=c(1,2))
)
run2 <- row.names(SeqTabNoChim) %in% c('E22C.fastq.gz', 'E23A.fastq.gz', 'E25A.fastq.gz',
                                       'E25B.fastq.gz', 'E25C.fastq.gz', 'E25D.fastq.gz',
                                       'E26A.fastq.gz', 'E26B.fastq.gz', 'E26C.fastq.gz',
                                       'E26D.fastq.gz', 'E27A.fastq.gz', 'E27B.fastq.gz',
                                       'E27C.fastq.gz', 'E28A.fastq.gz', 'E28B.fastq.gz',
                                       'E28C.fastq.gz', 'E28D.fastq.gz', 'E29A.fastq.gz',
                                       'E29B.fastq.gz', 'E29C.fastq.gz', 'E29D.fastq.gz',
                                       'E30A.fastq.gz', 'E30B.fastq.gz', 'E30C.fastq.gz',
                                       'E30D.fastq.gz', 'E31A.fastq.gz', 'E31B.fastq.gz',
                                       'E31C.fastq.gz', 'E31D.fastq.gz', 'E33A.fastq.gz',
                                       'E33B.fastq.gz', 'E33C.fastq.gz', 'E33D.fastq.gz',
                                       'E35A.fastq.gz', 'E35B.fastq.gz', 'E35C.fastq.gz',
                                       'E35D.fastq.gz', 'E36A.fastq.gz', 'E36B.fastq.gz',
                                       'E36C.fastq.gz', 'E36D.fastq.gz', 'E38A.fastq.gz',
                                       'E38B.fastq.gz', 'E38C.fastq.gz', 'E38D.fastq.gz',
                                       'E39A.fastq.gz', 'E39B.fastq.gz', 'E39C.fastq.gz',
                                       'E39D.fastq.gz', 'L23B.fastq.gz', 'L24C.fastq.gz',
                                       'L25A.fastq.gz', 'L25B.fastq.gz', 'L25C.fastq.gz',
                                       'L25D.fastq.gz', 'L26A.fastq.gz', 'L26B.fastq.gz',
                                       'L26C.fastq.gz', 'L26D.fastq.gz', 'L27A.fastq.gz',
                                       'L27B.fastq.gz', 'L27C.fastq.gz', 'L28A.fastq.gz',
                                       'L28B.fastq.gz', 'L28C.fastq.gz', 'L28D.fastq.gz',
                                       'L29A.fastq.gz', 'L29B.fastq.gz', 'L29C.fastq.gz',
                                       'L29D.fastq.gz', 'L30A.fastq.gz', 'L30B.fastq.gz',
                                       'L30C.fastq.gz', 'L30D.fastq.gz', 'L31A.fastq.gz',
                                       'L31B.fastq.gz', 'L31C.fastq.gz', 'L31D.fastq.gz',
                                       'L33A.fastq.gz', 'L33B.fastq.gz', 'L33C.fastq.gz',
                                       'L33D.fastq.gz', 'L35B.fastq.gz', 'L35C.fastq.gz',
                                       'L35D.fastq.gz', 'L36A.fastq.gz', 'L36B.fastq.gz',
                                       'L36C.fastq.gz', 'L36D.fastq.gz', 'L38A.fastq.gz',
                                       'L38B.fastq.gz', 'L38C.fastq.gz', 'L38D.fastq.gz',
                                       'L39A.fastq.gz', 'L39B.fastq.gz', 'L39C.fastq.gz',
                                       'L39D.fastq.gz')
SampleData$seqrun[run2] <- 2
row.names(SampleData) <- row.names(SeqTabNoChim)

isEarly <- SampleData$age == 'Early'
isLate  <- SampleData$age == 'Late'
EAr <- SeqTabNoChim[isEarly,]
LAr <- SeqTabNoChim[isLate,]
EarlyIsoline <- SampleData[isEarly, 'isoline']
LateIsoline  <- SampleData[isLate,  'isoline']
# I found it impossible to do the following with plyr:
EA <- do.call(rbind, lapply(split(as.data.frame(EAr), EarlyIsoline), colSums))
LA <- do.call(rbind, lapply(split(as.data.frame(LAr),  LateIsoline), colSums))
rm(EarlyIsoline, LateIsoline, isEarly, isLate, run2)
```

On `2020-02-27`, I also filtered amplicons (by abundance) before building the
phylogenetic tree. I realize now it was a mistake to use total abundance and
prevalence across early and late samples as filtering criteria. Clearly, amplicons
in early samples should be filtered independently from amplicons in late samples.

On `2020-04-13`, I used a very conservative filter for the redundancy analysis,
because two samples had very low sequence counts in one time point. Then, including
low abundance amplicons in the ordination would make the isolines where relative
abundances could not be measured that low stick out. That is, I had to represent
all isolines in the same space. Here, for the moment I just want to measure diversity.
I think diversity measures will be more accurate if I relax those filters and
let low abundance amplicons be counted. I can raise the bar later, if needed.

```{r filtering, warning=FALSE, message=FALSE}
# Per isoline:
EA.prop <- t(apply(EA, 1, function(x) x / sum(x)))
LA.prop <- t(apply(LA, 1, function(x) x / sum(x)))

# Abundance per replicate:
Ar <- SeqTabNoChim
rm(SeqTabNoChim)
Ar.prop <- t(apply(Ar, 1, function(x) x/sum(x)))

# Data frame of amplicon traits, available for use with picante package.
Traits <- data.frame(
  EA.meanProp = apply(EA.prop, 2, mean),
  LA.meanProp = apply(LA.prop, 2, mean),
  EA.prevalence = colSums(EA > 0),
  LA.prevalence = colSums(LA > 0)
)

MinProp <- 1.0e-6
MinPrev <- 3

p1 <- ggplot(Traits, aes(x=EA.meanProp, y=EA.prevalence)) +
  geom_point(size=0.5) + scale_x_log10() + ggtitle('Early') +
  geom_vline(xintercept=MinProp, color='red') + geom_hline(yintercept=MinPrev, color='red')
p2 <- ggplot(Traits, aes(x=LA.meanProp, y=LA.prevalence)) +
  geom_point(size=0.5) + scale_x_log10() + ggtitle('Late') +
  geom_vline(xintercept=MinProp, color='red') + geom_hline(yintercept=MinPrev, color='red')
grid.arrange(p1, p2, nrow=1)

EA.filter <- Traits$EA.meanProp >= MinProp & Traits$EA.prevalence >= MinPrev
LA.filter <- Traits$LA.meanProp >= MinProp & Traits$LA.prevalence >= MinPrev

rm(p1,p2,MinProp,MinPrev)
```

## Alignment

Aligning only filtered sequences:

```{r alignment, message=FALSE, warning=FALSE}
if (file.exists('alignment.RData')) {
   load('alignment.RData')
} else {
   AmpFilt <- sort(colnames(EA)[EA.filter | LA.filter])
   # AmpFilt <- unique(c(colnames(EA)[EA.filter], colnames(LA)[LA.filter]))  # same.
   alignment <- AlignSeqs(RNAStringSet(dna[AmpFilt]))
   BrowseSeqs(alignment, htmlFile='alignment.html', openURL=FALSE)
   save(alignment, file='alignment.RData')
   # Forces removal of tree upon update of alignment:
   if (file.exists('fitGTR.RData')) file.remove('fitGTR.RData')
}
```

## Phylogenetic tree

```{r tree, message=FALSE}
if (file.exists('fitGTR.RData')) {
   load('fitGTR.RData')
} else {
   library(phangorn)
   phang.align <- phyDat(as(DNAStringSet(alignment), "matrix"), type="DNA")
   dm <- dist.ml(phang.align)
   treeNJ <- NJ(dm) # Note, tip order != sequence order
   fit = pml(treeNJ, data=phang.align)

   fitGTR <- update(fit, k=4, inv=0.2)
   fitGTR <- optim.pml(fitGTR, model="GTR", optInv=TRUE, optGamma=TRUE,
                      rearrangement = "stochastic", control = pml.control(trace = 0))
   detach("package:phangorn", unload=TRUE) # avoids name clashes
   save(fitGTR, file='fitGTR.RData')
}
```

The tree object, of class `phylo`, is in `fitGTR$tree`. I am discovering the possibilities
of the `picante` package. I will add taxonomy information to the `Traits` data frame, in case
I use it later.

```{r traits}
Traits$Phylum  <- taxa[match(rownames(Traits), rownames(taxa)), 'Phylum']
Traits$Class   <- taxa[match(rownames(Traits), rownames(taxa)), 'Class']
Traits$Order   <- taxa[match(rownames(Traits), rownames(taxa)), 'Order']
Traits$Family  <- taxa[match(rownames(Traits), rownames(taxa)), 'Family']
Traits$Genus   <- taxa[match(rownames(Traits), rownames(taxa)), 'Genus']
Traits$Species <- taxa[match(rownames(Traits), rownames(taxa)), 'Species']
```

## Life history traits

The matrix `LH` contains the 10 original life-history trait variables, in their
original scales, with isoline number as row names. I will make standardized version
for later use, without modifying the original.

```{r LH, message=FALSE}
LHc <- apply(LH, 2, function(x) (x - mean(x, na.rm=TRUE)) / sd(x, na.rm=TRUE))
LHc.imp <- mice(LHc, m=10)
LH.all <- complete(LHc.imp, action='all')
validIsolines  <- row.names(EA)
validVariables <- c('CSearly', 'CSSlope', 'EarlyRS', 'Rsen', 'AvLF', 'ActuarialB')
LH.all <- lapply(LH.all, function(x, f1=validIsolines, f2=validVariables) x[f1, f2])
```

# Diversity

Diversity measures can be analysed by replicate, rather than by isoline. We use
better the information that way. But I need isoline-level measures of diversity
as well, to correlate with life-history traits.

I exclude low abundance and low prevalence amplicons, according to the filter
set up above. When working with individual replicates from both early and late
time

```{r diversity, fig.width=10}
## By isoline
#  ==========
#
# I make two d.f. and then bind them either by column or by row, to
# make the wide and long versions, for plotting.
Diversity.Early <- data.frame(
  isoline   = rownames(EA),
  PD.Early  = pd(EA[,EA.filter], fitGTR$tree, include.root = FALSE)[,1],
  SR.Early  = pd(EA[,EA.filter], fitGTR$tree, include.root = FALSE)[,2],
  Shannon.E = diversity(EA[,EA.filter], index='shannon'),
  Simpson.E = diversity(EA[,EA.filter], index='simpson'),
  SeqNum.E  = rowSums(EA[,EA.filter]),
  batch     = sapply(split(SampleData, SampleData$isoline),
                function(x) sum(x$seqrun == 2) / length(x$seqrun))
)

Diversity.Late <- data.frame(
  isoline   = rownames(LA),
  PD.Late   = pd(LA[,LA.filter], fitGTR$tree, include.root = FALSE)[,1],
  SR.Late   = pd(LA[,LA.filter], fitGTR$tree, include.root = FALSE)[,2],
  Shannon.L = diversity(LA[,LA.filter], index='shannon'),
  Simpson.L = diversity(LA[,LA.filter], index='simpson'),
  SeqNum.L  = rowSums(LA[,LA.filter]),
  batch     = sapply(split(SampleData, SampleData$isoline),
                function(x) sum(x$seqrun == 2) / length(x$seqrun))
)

DiverIso.wide <- cbind(Diversity.Early[,-7], Diversity.Late[,-1])
names(Diversity.Early) <- c('isoline','PD','SR','Shannon','Simpson','SeqNum','batch')
names(Diversity.Late)  <- c('isoline','PD','SR','Shannon','Simpson','SeqNum','batch')
Diversity.Early$age <- factor('Early', levels=c('Early','Late'))
Diversity.Late$age  <- factor('Late', levels=c('Early','Late'))
DiverIso.long <- rbind(Diversity.Early, Diversity.Late)

# I remove the SeqNum (sequencing effort) for sample 29, late, just for plotting.
z <- DiverIso.wide['29', 'SeqNum.L']
DiverIso.wide['29', 'SeqNum.L'] <- NA
pairs(DiverIso.wide[,c(2:11)])
DiverIso.wide['29', 'SeqNum.L'] <- z
rm(z, Diversity.Early, Diversity.Late)

## Per replicate
#  =============
DiverRep.long <- data.frame(
  isoline = SampleData$isoline,
  age     = SampleData$age,
  PD      = numeric(length = nrow(Ar)),
  SR      = numeric(length = nrow(Ar)),
  Shannon = numeric(length = nrow(Ar)),
  Simpson = numeric(length = nrow(Ar)),
  SeqNum  = numeric(length = nrow(Ar)),
  batch   = SampleData$seqrun
)
isEarly <- DiverRep.long$age %in% 'Early'
isLate  <- DiverRep.long$age %in% 'Late'
DiverRep.long[isEarly, 'PD'] <- pd(Ar[isEarly, EA.filter], fitGTR$tree, include.root=FALSE)[,1]
DiverRep.long[isLate,  'PD'] <- pd(Ar[isLate,  LA.filter], fitGTR$tree, include.root=FALSE)[,1]
DiverRep.long[isEarly, 'SR'] <- pd(Ar[isEarly, EA.filter], fitGTR$tree, include.root=FALSE)[,2]
DiverRep.long[isLate,  'SR'] <- pd(Ar[isLate,  LA.filter], fitGTR$tree, include.root=FALSE)[,2]
DiverRep.long[isEarly, 'Shannon'] <- diversity(Ar[isEarly, EA.filter], index='shannon')
DiverRep.long[isLate,  'Shannon'] <- diversity(Ar[isLate,  LA.filter], index='shannon')
DiverRep.long[isEarly, 'Simpson'] <- diversity(Ar[isEarly, EA.filter], index='simpson')
DiverRep.long[isLate,  'Simpson'] <- diversity(Ar[isLate,  LA.filter], index='simpson')
DiverRep.long[isEarly, 'SeqNum'] <- rowSums(Ar[isEarly, EA.filter])
DiverRep.long[isLate,  'SeqNum'] <- rowSums(Ar[isLate, EA.filter])
```

Species richness (`SR.Early` and `SR.Late`) is correlated with phylogenetic
distance (`PD.Early` and `PD.Late`), but not with sequencing effort (`SeqNum`).
Measures of diversity late in life are not visibly correlated with those early 
in life. 

```{r moreDiversity, fig.width=10}
p1 <- ggplot(DiverIso.long, aes(x=age, y=PD)) + geom_boxplot(notch=TRUE) + ylab('Phylogenetic Diversity')
p2 <- ggplot(DiverIso.long, aes(x=age, y=SR)) + geom_boxplot(notch=TRUE) + ylab('Species richness')
p3 <- ggplot(DiverIso.long, aes(x=age, y=Simpson)) + geom_boxplot(notch=TRUE) + ylab('Simpson index')
p4 <- ggplot(DiverIso.long, aes(x=age, y=Shannon)) + geom_boxplot(notch=TRUE) + ylab('Shannon index')
grid.arrange(p1,p2,p4, nrow=1)
rm(p1, p2, p3, p4)
```

The plot above shows roughly an effect of age on diversity, that Pau Carazo
had reported previously, using slightly different data (different amplicon
filters, and indices estimated with a different package). What we didn't know
is that the effect is so much more accused when using Faith's phylogenetic
diversity. In a model:

```{r diversityModel}
m1 <- lmer(PD ~ age + (age|isoline) + (1|batch), data=DiverRep.long)
summary(m1)
anova(m1)
```

This indicates that not only there are fewer *species* (amplicon variants) late
in life, but that the ones missing are quite diverse or spread phylogenetically.
This matches the results from the differential abundance analysis performed on
`2020-03-17`. You can see it [here](https://htmlpreview.github.io/?https://github.com/IgnasiLucas/fly/blob/master/results/2020-03-17/deseq2.html). In that report there is a figure with the fold-change of the amplicons
with significantly different abundance between early and late life. Amplicons are
distributed in the plot according to genus and color-coded according to phylum. There
are many more genera and phyla with negative than with positive change fold.

Among the amplicons with a significant fold change in `2020-03-17` there are
several *Acetobacter* and some *Lactobacillus*.

# Redundancy analysis

Much can be improved from the first redundancy analysis performed on `2020-04-13`.
Little variance in life-history traits could be explained by the main axes of
variation in amplicon abundance. A wiser selection of explanatory variables could
help compensate the noise and the high dimensionality in abundance data. I recall
Pau Carazo's multiple linear regression analyses were more successful than my
redundancy analysis. He found some effects of two ordination axes of abundance
data on two out of three principal components of life-history variable. Neither
diversity (Simpson index) nor *Lactobacillus* to *Acetobacter* abundance ratio
made any important contributions to Pau's models.

## Diversity and significant change

Here I will use phylogenetic diversity and abundances of only some of the amplicons
with the most significant fold change of abundance between early and late samples,
as determined in `2020-03-17`. From those results, data frame `sigtab` contains
the `r dim(sigtab)[1]` most significant amplicons, belonging to
`r length(unique(sigtab$Genus))` genera and `r length(unique(sigtab$Phylum))` phyla.


```{r rda1}
sigtab$amplicon <- rownames(sigtab)
# Need to remmove names attribute of variables in sigtab. Didn't find a way
# to avoid a loop!
for (var in colnames(sigtab)) {
  names(sigtab[,deparse(as.name(var))]) <- NULL
}
MaxDifPerGenus <- ddply(sigtab, .(Genus), function(x) head(x[order(x$padj),], n=1))
MaxDifPerGenus[,-c(7,10,13)]

# Using only late abundance for all amplicons, irrespective of fold change:
z1 <- LA.prop[, MaxDifPerGenus[MaxDifPerGenus$log2FoldChange < 0, 'amplicon']]
z2 <- LA.prop[, MaxDifPerGenus[MaxDifPerGenus$log2FoldChange > 0, 'amplicon']]
z1 <- z1[, colnames(z1) %in% colnames(LA)[LA.filter]]
z2 <- z2[, colnames(z2) %in% colnames(LA)[LA.filter]]
colnames(z1) <- paste0('L', colnames(z1))
colnames(z2) <- paste0('L', colnames(z2))
Selected <- cbind(z1, z2)

Predictors <- as.data.frame(log10(Selected + 1.0e-07))
Predictors$PD.Early <- DiverIso.wide$PD.Early
Predictors$PD.Late  <- DiverIso.wide$PD.Late
Predictors$Simpson.E <- DiverIso.wide$Simpson.E
Predictors$Simpson.L <- DiverIso.wide$Simpson.L
batch <- sapply(split(SampleData, SampleData$isoline),
                function(x) sum(x$seqrun == 2) / length(x$seqrun))
all.equal(names(batch), rownames(Predictors))

RDA0    <- lapply(LH.all, function(lh) rda(lh ~ Condition(batch),
                                        data = Predictors))
RDA18  <- lapply(LH.all, function(lh) rda(lh ~ . + Condition(batch),
                                           data = Predictors))
RDA18.anova <- lapply(RDA18, anova)
sapply(RDA18.anova, function(x) x$'Pr(>F)'[1])
#RDA  <- lapply(RDA0, function(x) ordistep(x, RDA18[[1]], Pin=0.1, Pout=0.15, steps=100, trace=FALSE))
#lapply(RDA, function(x) x$terms[[3]])
```

## Biodiversity and significant change, with PCA

```{r rda2}
# Use the amplicons in significant results (sigtab) as a filter for abundance data
F <- rownames(sigtab)
z1 <- EA.prop[,F]
z2 <- LA.prop[,F]
colnames(z1) <- paste0('E', colnames(z1))
colnames(z2) <- paste0('L', colnames(z1))
A <- cbind(z1,z2)
A.sq <- apply(A, 2, skewness)
B <- A[,abs(A.sq) <= 0.25]   # not transformed
C <- sqrt(A[,abs(A.sq) >  0.25 & abs(A.sq) <= 1.0])  # sqrt-transformed
D <- log10(A[,abs(A.sq) >  1.0] + 1.0e-07)   # log-transformed
colnames(C) <- paste0('sqrt.',colnames(C))
colnames(D) <- paste0('log.', colnames(D))
AT <- cbind(B,C,D)
AT <- AT[, abs(apply(AT,2,skewness)) < 2]
AT.cov <- cov.wt(AT)
AT.pca <- princomp(AT, covmat = AT.cov, scores = TRUE)
stopifnot(all.equal(rownames(AT.pca$scores), rownames(DiverIso.wide)))
Predictors <- cbind(DiverIso.wide[,c('PD.Early','PD.Late','Simpson.E','Simpson.L')],
                    AT.pca$scores[,1:17])
RDA0 <- lapply(LH.all, function(lh) rda(lh ~ Condition(batch),
                                        data = Predictors))
RDAX <- lapply(LH.all, function(lh) rda(lh ~ . + Condition(batch),
                                        data = Predictors))
RDAX.aov <- lapply(RDAX, anova)
sapply(RDAX.aov, function(x) x$'Pr(>F)'[1])
#RDA <- lapply(RDA0, function(x) ordistep(x, RDAX[[1]], Pin=0.1, Pout=0.15, steps=100, trace=FALSE))
#lapply(RDA, function(x) x$terms[[3]])
```

## Diversity and genus

Here I use as predictor variables the abundances of the most frequent genera either
early or late, and diversity measures.

```{r rda3}
z <- as.data.frame(t(EA.prop[,EA.filter | LA.filter]))
z$Genus <- taxa[rownames(z),'Genus']
sum(is.na(z$Genus))
GenPropE <- sapply(split(z[,1:24], z$Genus), function(x) colSums(x))
GeneraE <- names(sort(colSums(GenPropE), decreasing=TRUE)[1:19])

z <- as.data.frame(t(LA.prop[,EA.filter | LA.filter]))
z$Genus <- taxa[rownames(z), 'Genus']
sum(is.na(z$Genus))
GenPropL <- sapply(split(z[,1:24], z$Genus), function(x) colSums(x))
GeneraL <- names(sort(colSums(GenPropL), decreasing=TRUE)[1:19])

stopifnot(all.equal(rownames(GenPropL), rownames(DiverIso.wide)),
          all.equal(rownames(GenPropE), rownames(DiverIso.wide)))
PredictorsE <- as.data.frame(cbind(PD.Early = DiverIso.wide$PD.Early,
                                   Simpson.E = DiverIso.wide$Simpson.E,
                                   log10(GenPropE[,colnames(GenPropE) %in% GeneraE] + 1.0e-6)))
PredictorsE <- PredictorsE[, abs(sapply(PredictorsE, skewness)) < 1.5]
names(PredictorsE) <- paste0(names(PredictorsE), '.E')
PredictorsL <- as.data.frame(cbind(PD.Late = DiverIso.wide$PD.Late,
                                   Simpson.L = DiverIso.wide$Simpson.L,
                                   log10(GenPropL[,colnames(GenPropL) %in% GeneraL] + 1.0e-6)))
PredictorsL <- PredictorsL[, abs(sapply(PredictorsL, skewness)) < 1.5]
names(PredictorsL) <- paste0(names(PredictorsL), '.L')
RDA0 <- lapply(LH.all, function(lh) rda(lh ~ Condition(batch),
                                        data = PredictorsL))
RDAX <- lapply(LH.all, function(lh) rda(lh ~ . + Condition(batch),
                                        data = PredictorsL))
RDAX.anova <- lapply(RDAX, anova)
sapply(RDAX.anova, function(x) x$'Pr(>F)'[1])
RDA  <- lapply(RDA0, function(x) ordistep(x, RDAX[[1]], Pin=0.1, Pout=0.15, steps=100, trace=FALSE))
lapply(RDA, function(x) x$terms[[3]])


```

# Class

```{r rda4}
z <- as.data.frame(t(EA.prop[,EA.filter | LA.filter]))
z$Class <- taxa[rownames(z),'Class']
sum(is.na(z$Class))
ClassPropE <- sapply(split(z[,1:24], z$Class), function(x) colSums(x))
ClassE <- names(sort(colSums(ClassPropE), decreasing=TRUE))

z <- as.data.frame(t(LA.prop[,EA.filter | LA.filter]))
z$Class <- taxa[rownames(z), 'Class']
sum(is.na(z$Class))
ClassPropL <- sapply(split(z[,1:24], z$Class), function(x) colSums(x))
ClassL <- names(sort(colSums(ClassPropL), decreasing=TRUE)[1:19])

stopifnot(all.equal(rownames(ClassPropL), rownames(DiverIso.wide)),
          all.equal(rownames(ClassPropE), rownames(DiverIso.wide)))
PredictorsE <- as.data.frame(cbind(PD.Early = DiverIso.wide$PD.Early,
                                   Simpson.E = DiverIso.wide$Simpson.E,
                                   log10(ClassPropE[,colnames(ClassPropE) %in% ClassE] + 1.0e-7)))
PredictorsE <- PredictorsE[, abs(sapply(PredictorsE, skewness)) < 1.5]
names(PredictorsE) <- paste0(names(PredictorsE), '.E')
PredictorsL <- as.data.frame(cbind(PD.Late = DiverIso.wide$PD.Late,
                                   Simpson.L = DiverIso.wide$Simpson.L,
                                   log10(ClassPropL[,colnames(ClassPropL) %in% ClassL] + 1.0e-7)))
#PredictorsL <- PredictorsL[, abs(sapply(PredictorsL, skewness)) < 1.5]
names(PredictorsL) <- paste0(names(PredictorsL), '.L')
RDA0 <- lapply(LH.all, function(lh) rda(lh ~ Condition(batch),
                                        data = PredictorsL))
RDAX <- lapply(LH.all, function(lh) rda(lh ~ . + Condition(batch),
                                        data = PredictorsL))
RDAX.anova <- lapply(RDAX, anova)
sapply(RDAX.anova, function(x) x$'Pr(>F)'[1])
#RDA  <- lapply(RDA0, function(x) ordistep(x, RDAX[[1]], Pin=0.1, Pout=0.15, steps=100, trace=FALSE))
#lapply(RDA, function(x) x$terms[[3]])
```

## Strong abundance filter and diversity

```{r rda5, message=FALSE}
F <- Traits$EA.meanProp >= 1.0e-04 & Traits$LA.meanProp >= 1.0e-04
z1 <- EA.prop[,F]
z2 <- LA.prop[,F]
colnames(z1) <- paste0('E', colnames(z1))
colnames(z2) <- paste0('L', colnames(z2))
AF <- cbind(z1,z2)
AF.log <- log10(AF + 1.0e-07)
AF.log.n <- AF.log[, abs(apply(AF.log,2,skewness)) <= 1.5]
AF.log.cov <- cov.wt(AF.log.n)
AF.pca <- princomp(AF.log.n, covmat=AF.log.cov, scores=TRUE)
plot(AF.pca)
Predictors <- cbind(DiverIso.wide[,c('PD.Early','PD.Late','Simpson.E','Simpson.L')],
                    AF.pca$scores[,1:17])
RDA0 <- lapply(LH.all, function(lh) rda(lh ~ Condition(batch), data=Predictors))
RDAX <- lapply(LH.all, function(lh) rda(lh ~ . + Condition(batch), data=Predictors[,1:17]))
RDAX.aov <- lapply(RDAX, anova)
sapply(RDAX.aov, function(x) x$'Pr(>F)'[1])
RDA  <- lapply(RDA0, function(x) ordistep(x, RDAX[[1]], Pin=0.1, Pout=0.15, steps=100, trace=FALSE))
lapply(RDA, function(x) x$terms[[3]])
#RDA.aov <- lapply(RDA, anova)
#sapply(RDA.aov, function(x) x$'Pr(>F)'[1])
```

# Conclusions

I tried different ways to select the amplicon abundance-related variables that
could carry the most information about life history traits. Unfortunately, neither
diversity measures nor several summaries of abundance data produced significant models
at the global level. Before exploring further options, I want to summarize the attempts,
because, I admit, the code above is far from clear.

* First, PCA of roughly filtered relative abundances must contain too much noise: it is
impossible *a priori* to know what components may be related to life-history traits, and
an *ad hoc* selection of apparently significant components is bound to have false positive
results.
* Though phylogenetic diversity index is significantly different between early and late
samples, it does not help explain life-history variables.
* In an attempt to reduce the noise of the data, I added up abundances of amplicons of the
same genus (or class), following the intuition that taxa may be functionally redundant. But
taxa abundance did not improve our predictive power.
* I also used results from the differential abundance analysis, motivated by the idea that
amplicons that significantly change abundance from early to late life samples may be more
correlated with life history traits. For the moment, I used only a few amplicons (one from
each genera represented in the significant results), in an attempt to reduce the amount of
variables without using PCA. But it was not enough. I also used PCA, and nothing.

The global test always fails, and then I have to build a model by successive addition
or removal of terms. Even that only works if I relax the threshold p-values.

```{r sessionInfo}
save(SampleData, EA, LA, EA.prop, LA.prop, Ar, Ar.prop, Traits, EA.filter, LA.filter, fitGTR, DiverIso.wide, DiverIso.long, DiverRep.long, sigtab, taxa, file='Diversity.RData')
sessionInfo()
```

# References

