---
title: "Gut microbiota and life-history traits"
author: "J. Ignacio Lucas Lledó"
date: "27/4/2020"
output: html_document
bibliography: guts.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(vegan)
library(ggplot2)
library(gridExtra)
library(mice)          # for multiple imputation
library(e1071)         # for skewness calculation
library(ape)
library(picante)
library(ggtree)
library(tidytree)
library(GUniFrac)
load('../2020-04-22/Diversity.RData')
ls()
load('../2020-03-17/DESeq2.RData')
ls()
LH <- read.table('../../data/LifeHistoryVar.txt', header=TRUE, row.names=1)
```

# Amplicons or genera

When estimating and interpreting diversity, we can use amplicon sequence variants (ASV)
or genera. When using ASVs it is important to take sequence similarity into account,
because otherwise we could overestimate the functional diversity of groups with mostly
neutral diversity.

Matrices `Ar` and `Ar.prop` contain ASV counts and frequencies (respectively) in the
177 samples. I need to create equivalent matrices for genera, before filtering. There
are 82 amplicons without attributed genus. I will add the *Unknown* genus, to be able
to include those amplicons in the analysis.

```{r genera}
stopifnot(all.equal(rownames(Traits), colnames(Ar)))
Traits[is.na(Traits$Genus), 'Genus'] <- 'unknown'
# Having no NA values in Genus, I can use "==" below, instead of "%in%". 
# Using sapply(..., simplify=FALSE) instead of lapply() I get columns named by genus
Genera.counts <- do.call(cbind,
                         sapply(unique(Traits$Genus),
                                function(x, G=Traits, M=Ar) {
                                  f <- rownames(G[G$Genus == x,])
                                  rowSums(M[, f, drop=FALSE])
                                },
                                simplify = FALSE))
# I generate genus frequencies by adding up ASV frequencies, instead of normalizing
# the counts because the ASV relative frequencies may not add up to 1 if some ASVs
# have already been filtered out.
Genera.prop <- do.call(cbind,
                       sapply(unique(Traits$Genus),
                              function(x, G=Traits, M=Ar.prop) {
                                f <- rownames(G[G$Genus == x,])
                                rowSums(M[, f, drop = FALSE])
                              },
                              simplify = FALSE))
```

Here I am running most analyses keeping replicates separate, in contrast to
the analysis in `2020-04-22`, for example, where I joined the counts from
the 3 or 4 replicates of each isoline-age combination. This different approach
calls for a different filtering strategy. It is not appropriate to filter
ASVs by their isoline-level minimum prevalence or abundance, because at the
replicate level the power to detect ASVs may be 3 to 4 times lower. Thus, I
must be more demanding on the minimum proportion (too low frequency ASVs may
not be observable in individual samples) and on minimum prevalence (there are
only 24 isolines, but 177 samples). I also need to remove samples with too few
counts (E31D and L35C), before using them in the calculation of average
proportions and prevalences.

Filtering low frequency ASVs make samples with different sequencing effort more
comparable. A different question is if the same filter must be applied for early
and late samples. Up to now I have been using the same set of ASVs to measure
diversity early and late, by requiring the thresholds to be met among either early
or late samples. I realize this can have the effect of having very different
abundance distributions in early and late samples, especially if one category was
composed by a subset of the taxa in the other. Thus, this kind of filtering could
upwardly bias the difference in diversity between classes of samples.

For some applications below, I need all samples in the same matrix, and therefore
they must be scored for the abundance (or presence) of the same set of ASVs
(or genera). An alternative filtering strategy would
be to apply the filters globally: include ASVs (or genera) that meet the thresholds
over all samples, not only within their age class. I didn't use this approach before
because I expect it to have the effect of homogenizing the composition between
age classes: age class-specific ASVs would be less likely to meet the thresholds
than ASVs present in all ages. But I suppose it is better to be conservative when
testing for a difference in alpha diversity between early and late microbiomes.

```{r filters}
Sample.filter <- rowSums(Ar) >= 3000
Traits$meanProp <- apply(Ar.prop[Sample.filter,], 2, mean)
Traits$prevalence <- colSums(Ar[Sample.filter,] > 0)
MinProp <- 1.0e-5
MinPrev <- 5

ggplot(Traits, aes(x=meanProp, y=prevalence)) +
  geom_point(size=0.5) + scale_x_log10() +
  geom_vline(xintercept=MinProp, color='red') + geom_hline(yintercept=MinPrev, color='red')

ASV.filter <- Traits$meanProp >= MinProp & Traits$prevalence >= MinPrev

Genera.traits <- data.frame(
  meanProp = apply(Genera.prop[Sample.filter,], 2, mean),
  prevalence = colSums(Genera.prop[Sample.filter,] > 0)
)

ggplot(Genera.traits, aes(x=meanProp, y=prevalence)) +
  geom_point(size=0.5) + scale_x_log10() +
  geom_vline(xintercept=MinProp, color='red') + geom_hline(yintercept=MinPrev, color='red')

Genus.filter <- Genera.traits$meanProp >= MinProp & Genera.traits$prevalence >= MinPrev
```

# Alpha diversity

The change of $\alpha$ diversity with age is quite well resolved. I learned 
that amplicon diversity is not reduced with age. The reason is the high resolution
afforded by ASVs: even if one species (or genus) becomes dominant, it can
include a large number of ASVs (just as expected under the neutral theory),
to the point that ASV diversity remains high. Lesson 1: when using ASVs, a
more useful measure of diversity should incorporate the phylogenetic distances
among ASVs. To use the phylogenetic diversity index is the best option. An
alternative is to compute genus diversity, instead of ASV diversity. To illustrate
it, I include here a boxplot of phylogenetic diversities *per sample* (where samples
are not aggregated by isoline and age). This is quite redundant with `2020-04-22`.

```{r alpha}
ASV.alpha <- diversity(Ar[Sample.filter, ASV.filter], index = 'shannon')
boxplot(ASV.alpha ~ SampleData[Sample.filter, 'age'], xlab='Age', ylab='Shannon diversity',
        main = 'ASV diversity', log='y', notch=TRUE)
ASV.pd <- pd(Ar[Sample.filter, ASV.filter], fitGTR$tree, include.root=FALSE)[,1]
boxplot(ASV.pd ~ SampleData[Sample.filter, 'age'], xlab='Age', ylab='Phylogenetic diversity',
        main = 'ASV diversity', log='y',  notch=TRUE)
```

Let's compare it to the genus diversity.

```{r alphaGenus}
Genera.alpha <- diversity(Genera.counts[Sample.filter, Genus.filter], index = 'shannon')
boxplot(Genera.alpha ~ SampleData[Sample.filter, 'age'], xlab = 'Age', ylab = 'Shannon diversity',
        main = 'Genera diversity', log='y', notch=TRUE)
png(file='diversity.png', width = 480 * 3)
par(mfrow=c(1,3), cex=1.5)
boxplot(ASV.alpha ~ SampleData[Sample.filter, 'age'], xlab='Age', ylab='Shannon diversity',
        main = 'ASV diversity', log='y', notch=TRUE)
boxplot(ASV.pd ~ SampleData[Sample.filter, 'age'], xlab='Age', ylab='Phylogenetic diversity',
        main = 'ASV diversity', log='y',  notch=TRUE)
boxplot(Genera.alpha ~ SampleData[Sample.filter, 'age'], xlab = 'Age', ylab = 'Shannon diversity',
        main = 'Genera diversity', log='y', notch=TRUE)
dev.off()
```

Not only diversity is lower in late samples, but dispersion also seems lower.

# Beta diversity

## Test for homogeneity between early and late ages

How does microbiome change along life?  One possibility is that early in
life all isolines have a very similar and more generic microbiome, mostly
determined by food, which is common to all isolines. Then, over time,
isolines could diverge in the selective environments they offer to their
microbiomes, increasing diversity among isolines. Thus, even if alpha
diversity goes down, beta diversity would increase. Alternatively, the
colonization events may be quite variable among isolines (or even among
individual flies), in comparison to the final state late in life, as if
there was a convergent ecological succession in any fly. In that case,
not only alpha but also beta diversity would decrease.

In principle, I expect a decrease in $\beta$ diversity, because the general dominance
of *Acetobacter* amplicons in all late samples suggest a homogeneization of microbiota
with age. This can be tested with a *distance-based test for homogeneity
of multivariate dispersions* [@Anderson2006]. Below I use both `anova()` and
`permutest()` from the `vegan` package [@Oksanen2019].

To apply these tests we can use any distance measure between samples, because
they are turned into principal coordinates. If using simple $\beta$ diversity
metrics based on presence/absence data (those available in the `betadiver()`
function), a curious result emerges: only when using ASVs, and not if using
genera (or any other taxonomic level) the $\beta$ diversity is higher early
than late in life (not shown). That is quite counterintuitive, and after looking
deeper into it I realized the problem was in the distance metrics that do not
use quantitative data. The problem persists even if taking phylogeny into account
with the UniFrac distance, which also ignores the taxa abundances. There actually
is good reason to use other distance measures more popular among ecologists, such
as the Bray-Curtis dissimilarity. Because abundances are very skewed, Euclidean
distances are not appropriate. I do not normalize counts, but I do remove samples
E31D and L35C, with very low counts.

```{r betadiver}
BrayDist <- vegdist(Ar[Sample.filter, ASV.filter], method='bray')
stopifnot(all.equal(labels(BrayDist), rownames(SampleData[Sample.filter,])))
anova(betadisper(BrayDist, SampleData[Sample.filter, 'age']))
permutest(betadisper(BrayDist, SampleData[Sample.filter, 'age']))
plot(betadisper(BrayDist, SampleData[Sample.filter, 'age']))
boxplot(betadisper(BrayDist, SampleData[Sample.filter, 'age']), notch=TRUE, xlab='Age')
```

These results confirm that $beta$ diversity is higher early than late in life.
That is, the ASV turnover is higher early than late in life, among samples.
However, when dealing with $\alpha$ diversity, I dismissed diversity measures
of ASV data when not using phylogenetic information. Let's see what happens
when using phylogenetic information. I use the generalized UniFrac distances
[@Chen2012], which allow for different ways to weight abundance. It requires
the tree to be rooted. To root the tree I use as outgroup the small, monophyletic
group of Bacteroidetes.

```{r treedive}
Bacteroidetes <- fitGTR$tree$tip.label[fitGTR$tree$tip.label %in%
                                         rownames(Traits[Traits$Phylum=='Bacteroidetes',])]
RootedTree <- root(fitGTR$tree,
                   outgroup = Bacteroidetes,
                   resolve.root = TRUE)

unifracs <- GUniFrac(Ar[Sample.filter, ASV.filter], RootedTree, alpha=c(0, 0.5, 1))$unifracs
UFw <- as.dist(unifracs[, , 'd_1'])   # weighted UniFrac
UFu <- as.dist(unifracs[, , 'd_UW'])  # unweighted UniFrac
UFv <- as.dist(unifracs[, , 'd_VAW']) # variance-adjusted UniFrac
UF0 <- as.dist(unifracs[, , 'd_0'])   # GUniFrac with alpha 0
UF5 <- as.dist(unifracs[, , 'd_0.5']) # GUniFrac with alpha 0.5

stopifnot(all.equal(labels(UFw), rownames(SampleData[Sample.filter,])))
anova(betadisper(UFw, SampleData[Sample.filter, 'age'], add=TRUE))
permutest(betadisper(UFw, SampleData[Sample.filter, 'age'], add=TRUE))
plot(betadisper(UFw, SampleData[Sample.filter, 'age'], add=TRUE))
boxplot(betadisper(UFw, SampleData[Sample.filter, 'age'], add=TRUE),
        notch=TRUE, xlab='Age')

png(filename='beta.png', width=480*3)
par(mfrow=c(1,3), cex=1.2)
   plot(betadisper(UFu, SampleData[Sample.filter, 'age'], add=TRUE),
        sub='', main = 'Unweighted UniFrac')
   plot(betadisper(UF0, SampleData[Sample.filter, 'age'], add=TRUE),
        sub='', main = 'Generalised UniFrac, alpha = 0')
   plot(betadisper(UF5, SampleData[Sample.filter, 'age'], add=TRUE),
        sub='', main = 'Generalised UniFrac, alpha = 0.5')
par(mfrow=c(1,1))
dev.off()
```

## Test of homogeneity among isolines

The same method can be used to see if isolines (or isoline $\times$ age
combinations) have homogeneous dispersons in the ASV space. I will compare
isolines twice: first using early samples, and then late ones.

```{r interaction}
isEarly <- SampleData$age == 'Early'
isLate  <- SampleData$age == 'Late'
EarlyDist <- as.dist(GUniFrac(Ar[isEarly & Sample.filter, ASV.filter],
                              RootedTree, alpha=0)$unifracs[,,'d_0'])
LateDist  <- as.dist(GUniFrac(Ar[isLate & Sample.filter,  ASV.filter],
                              RootedTree, alpha=0)$unifracs[,,'d_0'])

anova(betadisper(EarlyDist, SampleData[isEarly & Sample.filter, 'isoline']))
permutest(betadisper(EarlyDist, SampleData[isEarly & Sample.filter, 'isoline']))
plot(betadisper(EarlyDist, SampleData[isEarly & Sample.filter, 'isoline']))
boxplot(betadisper(EarlyDist, SampleData[isEarly & Sample.filter, 'isoline']),
        xlab='Isoline', main='Early samples, Generalised UniFrac, alpha=0')

anova(betadisper(LateDist, SampleData[isLate & Sample.filter, 'isoline']))
permutest(betadisper(LateDist, SampleData[isLate & Sample.filter, 'isoline']))
plot(betadisper(LateDist, SampleData[isLate & Sample.filter, 'isoline']))
boxplot(betadisper(LateDist, SampleData[isLate & Sample.filter, 'isoline']),
        xlab='Isoline', main='Late samples, Generalised UniFrac, alpha=0')
```

The nice thing of the PCoA plots above is that I can see the variation
in composition among biological replicates, and among isolines. About the
boxplots, I remark first what we already know: early samples are further
away from their isoline centroids than late samples. The different amounts
of dispersion among isolines, both early and late in life are not significant,
probably due to the limited number of replicates (3 or 4).

## Analysis of variance using distance matrices

Another thing we can do with $\beta$ diversity is to see how well its
distribution corresponds to factors that characterize the samples, namely:
age, isoline and sequencing batch. The difference with the tests of
homogeneity performed above is that now I will be correlating the actual
position (not just the dispersions) of samples in the multidimensional
ASV space with their attributes. It is quite similar to a constrained
ordination analysis, like redundancy analysis. But in contrast to my
previous attempts to run redundancy analyses, here I focus in explaining
microbiome compositions, rather than using them to explain physiological
variation in the hosts.

```{r adonis2}
# No need to save the results
adonis2(UF0 ~ age * isoline + seqrun, data = SampleData[Sample.filter,], by='terms')
adonis2(UF5 ~ age * isoline + seqrun, data = SampleData[Sample.filter,], by='terms')
adonis2(Ar[Sample.filter, ASV.filter] ~ age * isoline + seqrun,
        data = SampleData[Sample.filter,], by='terms')
```

Results from different distance measures are quite equivalent: the isoline
and the specific way isolines respond to age are the main determinants of
the variation in microbiome composition. Age and sequencing batch are also
significant.

## Cannonical correspondance analysis

Exploratory ordination plots are available in `2020-03-25`. What I haven't
done yet is a constrained (or canonical) correspondance analysis (CCA) using age,
isoline and sequencing batch as constraining factors. It is probably unnecessary,
after having shown that both isoline and age, and their interaction, significantly
affect the ASV composition of microbiomes. 

```{r cca}
Ar.cca <- cca(Ar[Sample.filter, ASV.filter] ~ isoline + age + Condition(seqrun),
              data = SampleData[Sample.filter,])
plot(Ar.cca, display='bp')
```

Using ASV, instead of phyla or genera, prevents any useful display of *species* in the
biplot. Let's see if using genera we get something more interesting.

```{r genera2}
Genera.cca <- cca(Genera.counts[Sample.filter, Genus.filter] ~ isoline + age + Condition(seqrun),
                  data = SampleData[Sample.filter,])
Genera.cca
plot(Genera.cca, display='bp')
```

# Life history variables of isolines.

Life history variables include two missing data. I use 10 different imputations
to get a sense of the uncertainty due to missing data. Note I impute values
using all information available, and only then I select the isolines for which
we have microbiome data and the variables of interest.

```{r LH, message=FALSE}
LH$LateRS <- LH$EarlyRS - LH$Rsen
#LH <- LH[,-which(names(LH) %in% 'Rsen')]
LHc <- apply(LH, 2, function(x) (x - mean(x, na.rm=TRUE)) / sd(x, na.rm=TRUE))
set.seed(1007)
LHc.imp <- mice(LHc, m=10)
LH.all <- complete(LHc.imp, action='all')
validIsolines  <- row.names(EA)
validVariables <- c('CSearly', 'CSlate', 'EarlyRS', 'Rsen', 'AvLF', 'ActuarialB')
LH.all <- lapply(LH.all, function(x, f1=validIsolines, f2=validVariables) x[f1, f2])
```

Using life-history variables forces me to adopt
an isoline-centered perspective. I need to use now the `EA` and `LA` matrices
where abundance data from replicates is collapsed at the isoline level.

A first question is if diversity measures by themselves could explain any
of the variation in life history variables. I will use `DiverIso.wide` data
frame, created on `2020-04-22`, which includes measures of phylogenetic
diversity for early and late samples in every isoline. I add genus diversity.

```{r diversityRDA, warning=FALSE}
Genera.early <- do.call(rbind, lapply(split(as.data.frame(Genera.counts[isEarly,]),
                                            SampleData[isEarly, 'isoline']), colSums))
Genera.late  <- do.call(rbind, lapply(split(as.data.frame(Genera.counts[isLate, ]),
                                            SampleData[isLate,  'isoline']), colSums))
DiverIso.wide$Shannon.Genus.E <- diversity(Genera.early, index='shannon')
DiverIso.wide$Shannon.Genus.L <- diversity(Genera.late, index='shannon')
DiverIso.wide$Simpson.Genus.E <- diversity(Genera.early, index='simpson')
DiverIso.wide$Simpson.Genus.L <- diversity(Genera.late, index='simpson')

sapply(DiverIso.wide, skewness)
Predictors <- DiverIso.wide[, c('PD.Early','PD.Late','SR.Early','SR.Late',
                                'Shannon.E','Shannon.L','Shannon.Genus.E',
                                'Shannon.Genus.L')]

# First, I fit a null model with only the intercept
RDA0 <- lapply(LH.all, function(lh) rda(lh ~ 1, data=Predictors))
# This is the complete model, with all diversity measures
RDAX <- lapply(LH.all, function(lh) rda(lh ~ ., data=Predictors))
RDAX.aov <- lapply(RDAX, anova)
sapply(RDAX.aov, function(x) x$'Pr(>F)'[1])

# This is an automatic model selection procedure
RDA <- lapply(RDA0, function(x) ordiR2step(x, RDAX[[1]], trace=FALSE, Pin=0.1,
                                           Pout=0.15, R2scope=FALSE))
sapply(RDA, function(x) x$terms[[3]])
```

I don't see any measure of diversity, neither by itself nor combined with
others, significantly correlated with life history traits. I conclude that
having a more or less diverse microbiome does not, by itself, affect fitness.

# The role of amplicons or taxa that changed abundance between early and late time points

On `2020-03-17` I fit the model `~ seqrun + isoline + age` to amplicon
abundances using the `DESeq2` package. Amplicons with a significant
difference in abundance between early and late samples reflect a common
response from all or most isolines. When trying to explain the variance
in life-history traits among isolines, the most interesting amplicons
are actually those that vary among isolines.

Later on, I complemented the analysis in `2020-03-17`. I selected
the subset of amplicons for which a likelihood ratio test showed
that the *isoline* term produced a significantly better fit
(`AmpliconsIsolineTerm`); and the subset of them with a significant
interaction term between *isoline* and *age*.

I want to note that *DESeq2* does not allow for mixed models. Maybe I
could try the package *VariancePartition*, with the `dream()` function.

Below, I finally got a (marginally) significant global fit. It worked using
only the early abundances of the amplicons with the lowest pvalue
of the LRT for the interaction term, after transforming the relative
abundances to reduce the skew, and removing those that remained too
skewed. An appropriate maximum skewness proved essential to obtain the
almost significant results.

```{r interactionTerm}
SelectedAmplicons <- rownames(res2[res2$padj < 0.001, ])
#SelectedAmplicons  <- rownames(res2[order(res2$pvalue), ])[1:23]
z <- log10(EA.prop[, SelectedAmplicons] + 1.0e-07)
dim(z)
# I tried different skewness thresholds to get fewer than 24 ASVs.
z  <- z[, abs(apply(z, 2, skewness)) < 0.25]
dim(z)
if (dim(z)[2] > 23) {
  # Not used
  print('Using PCA')
  z.pca <- prcomp(z)
  Predictors <- as.data.frame(z.pca$x[, 1:22])
  print(sprintf("%% of variance: %.4f %%", 100 * sum(eigenvals(z.pca)[1:22])/sum(eigenvals(z.pca))))
} else {
  print(sprintf('Using %i amplicons', dim(z)[2]))
  Predictors <- as.data.frame(z[,1:(dim(z)[2])])
}
RDA0 <- lapply(LH.all, function(lh) rda(lh ~ 1, data=Predictors))
RDAX <- lapply(LH.all, function(lh) rda(lh ~ ., data=Predictors))
RDAX.aov <- lapply(RDAX, anova)
sapply(RDAX.aov, function(x) x$'Pr(>F)'[1])
```

The 10 values above are the p-values for the global test in the 10
datasets produced by multiple imputation. They are the lowest p values
I have seen in a global test till now. This is a good start to build
a model. Even relaxing the thresholds, a forward selection model departing
from the null model with only the intercept only adds one ASV (A0071,
*Acetobacter*, of course). Below I use a backward selection procedure
to remove from the complete model the irrelevant terms.

```{r ModelSelection}
RDA <- lapply(RDAX, function(x) ordistep(x, RDAX[[1]], trace=FALSE, Pin=0.05,
                                         direction = 'backward', Pout=0.1, steps=500))
#RDA <- lapply(RDA0, function(x) ordiR2step(x, RDAX[[1]], trace=FALSE,
#                                           Pin=0.15, Pout=0.2, R2scope=FALSE))
# Number of models a terms appears in:
TermUsage <- table(unlist(sapply(RDA, function(x) attr(x$terminfo$terms, 'term.labels'))))
TermUsage
```

Finally, I fit in all imputated datasets the same model with the
terms that got selected in at least 8 of those replicates.

```{r finalModel, fig.height=10, fig.width=10}
UsedTerms <- names(TermUsage[TermUsage >= 7])
RDA <- lapply(LH.all, function(lh) rda(lh ~ ., data=Predictors[,UsedTerms]))
sapply(RDA, function(x) summary(x)$constr.chi / summary(x)$tot.chi)
i <- 6
par(mfrow=c(2,2))
plot(RDA[[i]], choice=c(1,2), type='text', scaling=2,
     xlab=sprintf("RDA1 %.1f%%", 100 * summary(eigenvals(RDA[[1]]))[2,1]),
     ylab=sprintf("RDA2 %.1f%%", 100 * summary(eigenvals(RDA[[1]]))[2,2]))
plot(RDA[[i]], choice=c(3,2), type='text', scaling=2,
     xlab=sprintf("RDA3 %.1f%%", 100 * summary(eigenvals(RDA[[1]]))[2,3]),
     ylab=sprintf("RDA2 %.1f%%", 100 * summary(eigenvals(RDA[[1]]))[2,2]))
plot(RDA[[i]], choice=c(1,3), type='text', ylim=c(2, -2), scaling=2,
     xlab=sprintf("RDA1 %.1f%%", 100 * summary(eigenvals(RDA[[1]]))[2,1]),
     ylab=sprintf("RDA3 %.1f%%", 100 * summary(eigenvals(RDA[[1]]))[2,3]))
par(mfrow=c(1,1))

png(filename='RDA.png', width = 480*2, height = 480*2)
par(mfrow=c(2,2), cex=1.3)
plot(RDA[[1]], choice=c(1,2), type='text', scaling=2,
     xlab=sprintf("RDA1 (%.1f%%)", 100 * summary(eigenvals(RDA[[1]]))[2,1]),
     ylab=sprintf("RDA2 (%.1f%%)", 100 * summary(eigenvals(RDA[[1]]))[2,2]))
plot(RDA[[1]], choice=c(3,2), type='text', scaling=2,
     xlab=sprintf("RDA3 (%.1f%%)", 100 * summary(eigenvals(RDA[[1]]))[2,3]),
     ylab=sprintf("RDA2 (%.1f%%)", 100 * summary(eigenvals(RDA[[1]]))[2,2]))
plot(RDA[[1]], choice=c(1,3), type='text', ylim=c(2, -2), scaling=2,
     xlab=sprintf("RDA1 (%.1f%%)", 100 * summary(eigenvals(RDA[[1]]))[2,1]),
     ylab=sprintf("RDA3 (%.1f%%)", 100 * summary(eigenvals(RDA[[1]]))[2,3]))
dev.off()
```

There are two groups of ASVs with different,
apparently negative effects on fitness. First, A0065 and A0071
seem to *push* in a direction away from *CSlate* (late climbing speed).
I name this group "functional agers". Then, A0059, A0097 and A0123 move
away from *AvLF* (average life span, I think); I name this group "life
shorteners". Are those ASVs scattered in the *Acetobacter* phylogeny? Or are they
clusters of related clones? 

```{r tree, fig.width=10, fig.height=10}
TreeTraits <- Traits[names(fitGTR$data),]
TreeTraits$label <- rownames(TreeTraits)
TreeAceto <- drop.tip(fitGTR$tree, TreeTraits[! TreeTraits$Genus %in% 'Acetobacter', 'label'])
TibbAceto <- as_tibble(TreeAceto)
TibbAceto$modelTerms <- TibbAceto$label %in% UsedTerms
TDAceto <- as.treedata(TibbAceto)
TDAceto <- treeio::drop.tip(TDAceto, c('A1543', 'A1910', 'A2130', 'A2172',
                                      'A1818', 'A1067', 'A1306', 'A1787', 'A1962'))
ggtree(TDAceto, layout='circular') +
  geom_hilight(node=MRCA(TDAceto, c('A0059','A0064','A0097','A0123')), fill='blue', alpha=0.5) +
  geom_hilight(node=MRCA(TDAceto, c('A0065','A0071')), fill='red', alpha=0.5) +
  geom_cladelabel(node=MRCA(TDAceto, c('A0059','A0064','A0097','A0123')),
                  label='A0059, A0064, A0097, A0123',
                  offset=0.01, offset.text=0.2, hjust='center') +
  geom_cladelabel(node=MRCA(TDAceto, c('A0065','A0071')), label='A0065, A0071',
                  offset=0.01, offset.text=0.05, hjust='center')
ggsave('RDAtree.png')
```

They are clearly clustered, which suggests that several other ASV with related
sequences should also have similar effects.  

```{r test}
nodes.FA <- offspring(TDAceto, MRCA(TDAceto, c('A0065','A0071')), tiponly=TRUE)
nodes.LS <- offspring(TDAceto, MRCA(TDAceto, c('A0059','A0064','A0097','A0123')), tiponly=TRUE)
FunctionalAgers <- TibbAceto[TibbAceto$node %in% nodes.FA, 'label']$label
LifeShorteners  <- TibbAceto[TibbAceto$node %in% nodes.LS, 'label']$label

Predictors <- data.frame(
  FA = log10(apply(EA.prop[,FunctionalAgers], 1, mean) + 1.0e-7),
  LS = log10(apply(EA.prop[,LifeShorteners], 1, mean) + 1.0e-7)
)

RDA2 <- lapply(LH.all, function(lh) rda(lh ~ FA, data=Predictors))
RDA2.aov <- lapply(RDA2, anova)
sapply(RDA2.aov, function(x) x$'Pr(>F)'[1])
rm(RDA2.aov)
```

Either the combined or the average early abundances of presumed
*functional agers* or *life shorteners* added more noise than signal.
The relatinship with life history traits disappears when using all
ASV in their clusters. This does not invalidate the results, but
adds a note of caution.

# References


